---
subtitle: Trial Management Group Meeting, November 5, 2024
title: ADVANCE TRAUMA
format: pdf
---

```{r}
#| echo: false
#| include: false

# Source all functions
function.files <- list.files("../../functions",
    pattern = "\\.R$",
    full.names = TRUE
)
for (function.file in function.files) source(function.file)

# Attach global variables
attach(global_variables())
```

<!-- These notes were never circulated because they were compiled too late. -->

# Members Present

# Members Excused

# Agenda

1. **Update on trial progress**
   - Current status
   - Anticipated start date

2. **Plan for collecting adherence, quality of life, disability and return to work outcomes**
    <!-- - See @fig-nested-staircase. The number and frequency of shifts per month will be decided based on the number of patients per cluster and month. -->

```{r nested-staircase}
#| echo: false
#| label: fig-nested-staircase
#| fig.cap: "Nested staircase design for collecting adherence, quality of life, disability and return to work outcomes. Lines represent the duration of patient enrolment across clusters and phases. Data collection for these outcomes will happen three months before and three months after the training. During these months, we need to include at least 4 patients per cluster and month for adherence and 6 patients per cluster and month for quality of life, disability and return to work outcomes."
#| fig.align: "center"
#| message: FALSE

knitr::include_graphics(
    create_trial_design_flowchart(
        clusters = clusters, sequences = sequences, batches = batches,
        min.standard.care.months = 3,
        min.intervention.months = 3,
        batches.overlap.months = 3,
        start.month = 1,
        transition.months = transition.months,
        transition.overlap.months = transition.overlap.months,
        staircase.months = 3,
        return.figure = FALSE,
        device = "png"
    )
)
```

3. **Discussion on batch 2 sites**
   - Current candidate sites:
      - Nilratan Sircar Medical College & Hospital, Kolkata
      - Dayanand Medical College & Hospital, Ludhiana
      - PGIMER, Chandigarh?
      - Sion Hospital, Mumbai?
      - KEM Hospital, Mumbai?
      - AIIMS Bhubaneshwar can be tapped, if necessary.

4. **Additional points**

# Notes and Decisions

1. Prashant presented the progress with each site. 
2. We decided to collect adherence, quality of life, disability and return to work outcomes using the nested staircase design. This means that the sites will be informed about their implementation sequence three months before the start of the intervention. This was agreed to be acceptable. 
3. We discussed the candidate sites for the second batch. 