---
title: "Statistical Analysis Plan"
subtitle: | 
  | ADVANCE TRAUMA
  |
  | Effects of Advanced Trauma Life Support^®^ Training Compared to Standard Care on Adult Trauma Patient Outcomes: A Cluster Randomised Trial
author: "{{< var version_type >}} {{< var version >}}, {{< var date >}}"
bibliography: ../shared-assets/bibliography.bib
csl: nature.csl
---

```{r setup}
#| echo: FALSE

noacsr::source_all_functions()
attach(global_variables())
```

{{< pagebreak >}}

# Administrative information

## Study identifiers

```{r}
#| echo: FALSE

protocol.variables <- yaml::read_yaml("../protocol/_variables.yml")
```

-   Protocol version `r protocol.variables$version` dated `r protocol.variables$date`

{{< include ../shared-assets/study-identifiers.qmd >}}

## Changelog

Once version 1.0.0 is finalised, this section will be updated with a changelog.

<!--

| Version | Date       | Details | 
|---------|------------|---------|

-->

## Contributors

| Name and ORCID                                                                  | Affiliation                              | Role                       |
|---------------------------------------------------------------------------------|------------------------------------------|----------------------------|
| Martin Gerdin Wärnberg (MGW) `r insert_orcid_link("0000-0001-6069-4794")`       | Karolinska Institutet                    | Principal Investigator     |
| Anna Olofsson (AO) `r insert_orcid_link("0000-0002-9460-108X")`                 | Karolinska Institutet                    | Statistician, TMG member   |
| Karla Hemming (KH) `r insert_orcid_link("0000-0002-2226-6550")`                 | University of Birmingham, Birmingham, UK | Statistician, TMG member   |
| James Martin (JM) `r insert_orcid_link("0000-0002-6949-4200")`                  | University of Birmingham, Birmingham, UK | Statistician               |
| Jessica Kasza `r insert_orcid_link("0000-0002-8940-0136")`                      | Monash University, Melbourne, Australia | Statistician               |

{{< pagebreak >}}

# Trial synopsis

{{< include ../shared-assets/synopsis.qmd >}}

{{< pagebreak >}}

# Special considerations

{{< include ../shared-assets/special-considerations.qmd >}}

{{< pagebreak >}}

# Aim and objectives

## Aim

To compare the effects of ATLS^®^ training with standard care on outcomes in adult trauma patients.

## Objectives

- To estimate the effects of ATLS^®^ training on 30-day in-hospital mortality compared to standard care.
- To assess the robustness of the main analysis results to different model specifications through sensitivity analyses.
- To estimate the effects of ATLS^®^ training on secondary outcomes compared to standard care.
- To explore the effects of ATLS^®^ training on subgroups of patients compared to standard care.

# Statistical analysis

## Design

This is a batched stepped-wedge cluster randomised trial, including a total of `r clusters` clusters (see @fig-trial-design). The trial will be composed of `r batches` batches of identical `r total.months - 1`-period `r sequences`-sequence designs, with one cluster being assigned to each sequence of each batch [@Kasza2022]. Each period is one month, and each cluster will be in the trial for a total of `r total.months` months. The intervention will be implemented during a one-month transition period, which will be excluded from the analysis. There will be an overlap of `r batches.overlap.months` months between successive batches.

```{r trial-design}
#| echo: false
#| label: fig-trial-design
#| fig.cap: "Trial design. Lines represent the duration of patient enrolment across clusters and phases. Clusters will be sequentially allocated to a batch based on when they enter the study. Within each batch clusters will then be randomised to an intervention implementation sequence."
#| fig.align: "center"
#| message: FALSE
#| warning: FALSE

knitr::include_graphics(
    create_trial_design_flowchart(
        clusters = clusters, sequences = sequences, batches = batches,
        min.standard.care.months = min.standard.care.months,
        min.intervention.months = min.intervention.months,
        batches.overlap.months = batches.overlap.months,
        transition.months = transition.months,
        transition.overlap.months = transition.overlap.months,
        return.figure = FALSE,
        device = "png"
    )
)
```

Because certain secondary outcomes will be more burdensome to collect and because we anticipate larger effects on these outcomes, we will nest a staircase design within the main stepped-wedge design to measure these outcomes (see @fig-nested-staircase-design) [@grantham_staircase_2024]. The staircase design will include a random subset of patients presenting during the three months preceding the transition phase, and the three months following the transition phase.  

```{r nested-staircase-design}
#| echo: false
#| label: fig-nested-staircase-design
#| fig.cap: "Nested staircase design. The design includes a random subset of patients presenting during the three months preceding the transition phase, and the three months following the transition phase."
#| fig.align: "center"
#| message: FALSE
#| warning: FALSE

knitr::include_graphics(
  create_trial_design_flowchart(
    clusters = clusters, sequences = sequences, batches = batches,
    min.standard.care.months = min.standard.care.months,
    min.intervention.months = min.intervention.months,
    batches.overlap.months = batches.overlap.months,
    transition.months = transition.months,
    transition.overlap.months = transition.overlap.months,
    staircase.months = staircase.months,
    return.figure = FALSE,
    device = "png"
  )
)
```

## Randomisation of clusters

<!-- Need to add how clusters will be randomised -->

We will assign clusters to batches as they are found to be eligible and receive ethical approval. Batches will include clusters from hospitals in different regions to optimize trial logistics. We will randomise the clusters allocated to each batch to the different intervention implementation sequences within that batch. We will balance the randomisation within each batch on cluster size, defined as monthly volume of eligible patient participants, using covariate constrained randomisation. The random allocation sequence will be generated by JM who will send the results to the trial team, to allow the logistics of securing seats in upcoming courses to be initiated. We will conceal the randomisation order to the sites for as long as it is logistically possible, considering that arrangements at the site level for sending physicians to ATLS^®^ training need to be made in advance.
 
 <!--
[^randomisation]: **TODO:** James gets back with a short snippet with more details on how the randomisation will be done.
-->

## Outcomes

{{< include ../shared-assets/outcomes.qmd >}}

## Statistical hypotheses

Our primary statistical hypotheses are:

-   **Null hypothesis**: There is no difference in the primary outcome of 30-day in-hospital mortality between those randomised to ATLS^®^ and standard care, meaning that the odds ratio (OR) for ATLS^®^ vs standard care would be 1.
-   **Alternative hypothesis**: There is an absolute difference in the primary outcome of 30-day in-hospital mortality between those randomised to ATLS^®^ and standard care of at least 5% units, meaning that the OR for ATLS^®^ vs standard care would be different from 1. Our expectation, based on our pilot study and review of the literature, is that the OR will be less than 1, indicating lower odds of 30-day in-hospital mortality among those randomised to ATLS^®^ group compared to those randomised to the standard care group.

## Sample size calculations

{{< include ../shared-assets/sample-size-calculations.qmd >}}

## Statistical principles

### Statistical software

We will use the R Statistical Software for all analyses [@R].

### Levels of statistical significance and confidence

We will use a two-sided significance level of 0.05 for all analyses, and we will report 95% confidence intervals (CI) for all estimates. We will not adjust for multiple testing because no secondary outcome is regarded as singularly more important.

## Analysis populations

The unit of randomisation is the hospital, because all units trained in the same hospital will be treated as one cluster, but the unit of analysis is the individual patient. The group allocation for a patient depends on the period in which the patient was admitted to the hospital, and patients will be considered exposed to the intervention if they were admitted to the hospital at any time point following the transition period. We will use an intention-to-treat approach for all analyses, according to the planned date of transition from standard care. We might consider a per protocol analysis if clusters deviate considerably from the planned date of transition, using the actual date of transition as the cut-off point. We will report the results of the per protocol analysis as a sensitivity analysis. We will use a CONSORT diagram to display the flow of hospitals, clusters and patients through the trial. We will present cluster level summaries of the intervention effect. We will report the study according to the CONSORT guidelines for stepped-wedge randomised trials [@Hemming2018].

## Baseline analyses

### Cluster characteristics

We will describe cluster characteristics including location and size using frequencies and percentages for discrete variables and means, standard deviations, medians and interquartile ranges (Q1-Q3) for continuous variables.

### Patient characteristics

We will describe patient characteristics at baseline, meaning all pre-training periods, per treatment group and overall using frequencies and percentages for discrete variables and means, standard deviations, medians and interquartile ranges (Q1-Q3) for continuous variables. We will not adjust for clustering when presenting baseline characteristics.

## Interim analysis

{{< include ../shared-assets/interim-analysis.qmd >}}

<!-- 
Removed the recalculation of the sample size from the interim analysis after discussion with Karla and James on 2024-10-16. The rationale was that the estimates would be associated with so much uncertainty and that it wouldn't change things.

We will recalculate the sample size based on the observed period sizes and the observed effect size, using the Shiny CRT Calculator [@Hemming2020Feb]. In the revised sample size calculation, we will assume an ICC of 0.02, but consider sensitivity across the range 0.01-0.05 [@Campbell2005; @Eldridge2015], and a CAC of 0.9, but consider sensitivity across the range 0.8-1.0 [@Hemming2020Feb; @Martin2016; @Korevaar2021]. These assumptions are the same as in the original sample size calculation. -->

## Analysis of the primary outcome

The primary outcome is in-hospital mortality within 30 days of arrival at the emergency department and will be analysed as a dichotomous variable.

We will estimate the intervention effect as the odds ratio (OR) of death between the ATLS® and standard care arms, with an OR < 1 indicating lower odds of death in the ATLS® arm compared to the standard care arm and vice versa. 

We will also estimate the absolute risk difference (ARD) of death between the ATLS® and standard care arms, with an ARD < 0 indicating lower risk of death in the ATLS® arm compared to the standard care arm and vice versa.

We will use a mixed-effects binomial regression model to estimate the effect of ATLS® on mortality. The OR will be estimated using the logit link, and the ARD will be estimated using the identity link.

The primary analysis will be based on the most complex model that converges from a prespecified sequence of models of decreasing complexity, and this model will then be subject to sensitivity analyses. 

The analysis will proceed from the most complex model to progressively simpler models if convergence or numerical stability issues arise. If the binomial model with the logit link converges but the corresponding model with the identity link does not, then we will report only an OR. 

All models in this sequence will include a fixed effect for intervention exposure and adjust for time, although the parametrisation of time and the random effects structure may be simplified.

The model sequence is defined as follows:

1. **Primary model**: a model including fixed effects for period, modelled as a categorical variable with separate period effects for each batch, and a fixed effect for intervention exposure. Clustering will be accounted for by including a random intercept for cluster and a random cluster-by-period effect, both assumed to follow a normal distribution. The model is specified in @eq-primary-model.

$$
g\{\Pr(Y_{bkti} = 1)\} = μ + β_bt + θ X_bkt + α_bk + γ_bkt
$$
{#eq-primary-model}

2. **Simplified random effects structure**: a model including fixed effects for period and intervention exposure and a random intercept for cluster only, thereby removing the random cluster-by-period effect. The model is specified in @eq-simplified-random-effects-model.

$$
g\{\Pr(Y_{bkti} = 1)\} = μ + β_bt + θ X_bkt + α_bk
$$
{#eq-simplified-random-effects-model}

3. **Model with shared period effects**: a model in which period effects are shared across batches, with time adjusted for using a common period effect or calendar time modelled directly (e.g., as a continuous covariate or using splines), while retaining the random effects structure. The model is specified in @eq-shared-period-model.

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \beta_{t} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt}
$$
{#eq-shared-period-model}

4. **Combined simplified model**: a model that includes both a simplified random effects structure (random intercept for cluster only) and shared period effects. This represents the most parsimonious mixed-effects model considered acceptable for the primary analysis. The model is specified in @eq-combined-simplified-model.

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \beta_{t} + \theta X_{bkt} + \alpha_{bk}
$$
{#eq-combined-simplified-model}

Where:

- $g(\cdot)$: the link function, specified as either the logit link,
$g(p) = \log\{p/(1-p)\}$, or the identity link, $g(p) = p$,
depending on the estimand of interest.
- $\Pr(Y_{bkti} = 1)$: the probability of death for patient $i = 1, \dotsc, m$ in cluster $k = 1, \dotsc, \texttt{r clusters}$ during period $t = 1, \dotsc, \texttt{r total.months - 1}$ in batch $b = 1, \dotsc, \texttt{r batches}$.
- $\mu$: the intercept, representing the value of the linear predictor for the reference levels of all fixed effects, with random effects equal to zero.
- $\beta_t$: the fixed effect of period $t$, shared across batches.
- $\beta_{bt}$: the fixed effect of period $t$ in batch $b$, accounting for a separate period effect for each batch, resulting in a total of $\texttt{r batches} \times (\texttt{total.months} - 1)$ period effects.
- $\theta$: the fixed effect of intervention exposure; under the logit link this represents the log-odds ratio for death associated with ATLS® exposure compared with standard care, and under the identity link it represents the absolute risk difference.
- $X_{bkt}$: the intervention exposure indicator for cluster $k$ during period $t$ in batch $b$, with $X_{bkt} = 1$ for ATLS® and $X_{bkt} = 0$ for standard care.
- $\alpha_{bk}$: a normally distributed random effect for cluster $k$ in batch $b$, representing between-cluster (hospital-level) variability in the outcome.
- $\gamma_{bkt}$: a normally distributed random effect for cluster $k$ in period $t$ in batch $b$, representing within-cluster variation over time and capturing cluster-specific deviations from the average period effect.

Where:

- $g(\cdot)$: the link function, specified as either the logit link, $g(p) = \log\{p/(1-p)\}$, or the identity link, $g(p) = p$, depending on the estimand of interest.
- $\theta$: **the intervention effect**, represented as a fixed effect of intervention exposure (the primary estimand); under the logit link this represents the log-odds ratio for death associated with ATLS® exposure compared with standard care, and under the identity link it represents the absolute risk difference.
- $\Pr(Y_{bkti} = 1)$: the probability of death for patient $i = 1, \dotsc, m$ in cluster $k = 1, \dotsc, \texttt{r clusters}$ during period $t = 1, \dotsc, \texttt{r total.months - 1}$ in batch $b = 1, \dotsc, \texttt{r batches}$.
- $\mu$: the intercept, representing the value of the linear predictor for the reference levels of all fixed effects, with random effects equal to zero.
- $\beta_t$: the fixed effect of period $t$, shared across batches.
- $\beta_{bt}$: the fixed effect of period $t$ in batch $b$, accounting for a separate period effect for each batch, resulting in a total of $\texttt{r batches} \times (\texttt{total.months} - 1)$ period effects.
- $X_{bkt}$: the intervention exposure indicator for cluster $k$ during period $t$ in batch $b$, with $X_{bkt} = 1$ for ATLS® and $X_{bkt} = 0$ for standard care.
- $\alpha_{bk}$: a normally distributed random effect for cluster $k$ in batch $b$, representing between-cluster (hospital-level) variability in the outcome.
- $\gamma_{bkt}$: a normally distributed random effect for cluster $k$ in period $t$ in batch $b$, representing within-cluster variation over time and capturing cluster-specific deviations from the average period effect.

The distributional assumptions for the random effects are specified in @eq-distributional-assumptions.

$$
\begin{aligned}
\alpha_{bk} &\overset{\text{iid}}{\sim} \mathcal{N}(0,\sigma^2_{\alpha}), \\
\gamma_{bkt} &\overset{\text{iid}}{\sim} \mathcal{N}(0,\sigma^2_{\gamma}), \\
\alpha_{bk} &\perp \gamma_{bkt}.
\end{aligned}
$$
{#eq-distributional-assumptions}

To correct potential inflation of the type I error rate due to the small number of clusters, a small-sample correction will be applied. Our primary method will be the Kenward–Roger correction [@grantham_evaluating_2022], but this choice may be updated based on methodological developments available closer to the end of patient inclusion. The correction applied may also differ for outcomes collected via the complete stepped-wedge design and the incomplete nested staircase designs.

All models will be fitted using residual pseudo-likelihood estimation based on linearisation with subject-specific expansion (RSPL), due to its computational stability and robust performance in small-sample settings. If the binomial model with the logit link converges but the corresponding model with the identity link does not, then we will report only an OR. 

<!--
*Question:** Should we adjust for calendar time because the last batch will be two years later than the first batch?

JK: Would this be instead of the beta_bt terms? That is, instead of separate period effects for each batch, just terms for each period would be included? (This is related to the question about 78 parameters in the model). The separate period effects for each batch will adjust for the calendar time. This is the model that makes the fewest assumptions about the period effects. 

We've done a bit of work recently which shows that if there are separate period effects in each batch but instead shared period effects are included, this can lead to quite a lot of bias in the estimation of the treatment effect. This work also seems to indicate that when separate period effects are included but the period effects are actually shared, this doesn't really have any deleterious impacts on the inference for the treatment effect. (However note that parameterisation of time wasn't the focus of the work we've done).

So my recommendation would be for the primary analysis to be as currently specified (separate period effects for each batch); then if there are issues with fitting that model, try a model with calendar time (i.e. period effects shared when there is overlap between successive batches).

**Question:** Just a thought: keep in mind the complexity of the random effects structure; if it leads to convergence problems or overfitting, should we consider simplifying the random effects?

-->

### Sensitivity analyses

We will conduct sensitivity analyses to assess the robustness of the primary analysis results to different modelling assumptions. In all sensitivity analyses, the focus will be on the stability of the estimated intervention effect $\theta$. Unless otherwise stated, each sensitivity analysis will be operationalised using two separate models, one with the logit link and one with the identity link.

#### Models with autoregressive within-cluster correlation

We will explore whether models with more complex correlation structures provide a better fit to the data. These models are not used as primary analysis models because there is limited understanding of when such models converge and how best to choose between different plausible correlation structures. Instead, we include these models as sensitivity analyses.

We will first explore whether allowing temporal correlation of outcomes within clusters affects inference on the intervention effect. Specifically, we will consider a discrete time-decay correlation structure by specifying a random cluster effect with an autoregressive structure of order one (AR(1)), described in @eq-ar1. This parameterisation assumes that correlations between cluster-level random effects decay geometrically with increasing temporal separation between periods.

$$
\alpha_{bk,t} = \rho \alpha_{bk,t-1} + \epsilon_{t}, \quad \epsilon_{t} \sim \mathcal{N}(0, \sigma^2_\alpha)
$$
{#eq-ar1}

Where:

- $\rho$ is the correlation parameter governing the association between adjacent periods within the same cluster.
- $\alpha_{bk,t}$ is the random effect for cluster $k$ in period $t$ in batch $b$.
- $\epsilon_t$ is a normally distributed error term.

We note that this AR(1) formulation differs slightly from alternative parameterisations used elsewhere in which correlations are specified through a joint covariance matrix. Although both approaches capture decaying temporal correlation, the parameter $\rho$ is not directly equivalent to parameters used in other specifications and should not be interpreted interchangeably.

#### Models with batch-specific secular trends

To assess sensitivity to assumptions about common secular trends across batches, we will include a random batch-by-period effect allowing each batch to follow its own time trend. This specification accounts for the randomisation by batches and permits additional between-batch heterogeneity over time. The model is specified in @eq-model-batch-trend.

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \beta_{bt} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt} + \delta_{bt}
$$
{#eq-model-batch-trend}

Where $\delta_{bt}$ represents a random effect for batch $b$ during period $t$, capturing between-batch variation in secular trends over time.


<!-- 
Need to be updated
```{r}
#| echo: FALSE
#| include: FALSE
#| eval: FALSE

library(nlme)

# Assuming `data` is your DataFrame with the necessary columns
# Fit the model
model <- nmle::gls(
    fixed = outcome ~ period + intervention + batch,
    data = data,
    correlation = corAR1(form = ~ 1 | cluster_id),
    weights = varIdent(form = ~ 1 | period)
)

# Check the summary
summary(model)
```
-->

#### Models with random cluster by intervention effects

To assess sensitivity to the assumption of a common intervention effect across clusters, we will extend the primary analysis model to allow the intervention effect to vary by cluster. Specifically, we will include a random cluster-level slope for the intervention exposure. The model is specified in @eq-model-random-cluster-intervention.

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \beta_{bt} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt} + u_{bk} X_{bkt}
$$
{#eq-model-random-cluster-intervention}

Where:

- $u_{bk} \sim \mathcal{N}(0,\sigma_u^2)$ is a random effect representing cluster-specific deviations from the average intervention effect.
- $u_{bk}$ is assumed to be independent of the cluster random effect $\alpha_{bk}$ and the cluster-by-period random effect $\gamma_{bkt}$.

#### Models with correlated random cluster-by-intervention effects

We will additionally explore sensitivity to the assumption that the random cluster-by-intervention effect is independent of the other random effects in the model. In this sensitivity analysis, we will fit a model that allows the random cluster-by-intervention effect $u_{bk}$ to be correlated with both the cluster random intercept $\alpha_{bk}$ and the cluster-by-period random effect $\gamma_{bkt}$. This relaxes the independence assumptions of the random-effects structure and allows for potential associations between baseline cluster risk, cluster-specific time deviations, and cluster-specific intervention effects. The model is specified in @eq-model-correlated-random-effects.

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \beta_{bt} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt} + u_{bk} X_{bkt}
$$
{#eq-model-correlated-random-effects}

We assume a joint multivariate normal distribution for the random effects:

$$
\begin{pmatrix}
\alpha_{bk} \\
\gamma_{bkt} \\
u_{bk}
\end{pmatrix}
\sim
\mathcal{N}\!\left(
\begin{pmatrix}
0 \\
0 \\
0
\end{pmatrix},
\begin{pmatrix}
\sigma^2_{\alpha} & \sigma_{\alpha\gamma} & \sigma_{\alpha u} \\
\sigma_{\alpha\gamma} & \sigma^2_{\gamma} & \sigma_{\gamma u} \\
\sigma_{\alpha u} & \sigma_{\gamma u} & \sigma^2_{u}
\end{pmatrix}
\right).
$$

#### Models with time modelled with a spline function

We will further explore the potential for a time-varying treatment effect [@kenny_analysis_2022]. To explore if the fixed period effect is both parsimonious and adequate to represent the extent of any underlying secular trend, we will model calendar time using a shared natural cubic spline basis across all batches with knots at the equally spaced time points 3, 6 and 9. This will result in five spline basis functions, because the natural cubic splines are modelled with three degrees of freedom but are constrained to be linear before the first and after the last knot. The model is specified in @eq-spline-model.

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \sum_{j=1}^{5} \beta_j S_j(t, \{3,6,9\}) + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt}
$$
{#eq-spline-model}

Where:

- $t$ denotes calendar time (period index) common to all batches.
- $S_j(t, \{3,6,9\})$ denotes the $j$th natural cubic spline basis function, shared across batches.
- $\beta_j$ is the coefficient for the $j$th spline basis function.

#### Models exploring lag and weaning effects

We will also extend the primary analysis model to include an interaction between treatment and number of periods since first treated, to examine if there is any indication of a relationship between duration of exposure to the intervention and outcomes. This will allow us to model different lag effects (whereby it takes time for the intervention to become embedded within the culture before its impact can properly start to be realised); as well as weaning effects (whereby the effect of the intervention starts to decrease – or fade). This type of analysis attempts to disentangle how some clusters end up having a long exposure to the intervention and others have a much shorter exposure time. The model is specified in @eq-lag-weaning-model.

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \beta_{bt} + \theta X_{bkt}
+ \theta_{\text{int}} X_{bkt} \times T_{bkt}
+ \alpha_{bk} + \gamma_{bkt}
$$
{#eq-lag-weaning-model}

Where $\theta_{\text{int}}$ represents the interaction between intervention exposure and time since first treated, and $T_{bkt}$ is the number of periods since first exposure.

### Adjusted analyses

Fully adjusted covariate analysis will additionally adjust for[^additional-covariates] (expected % missing data in parentheses):

-   Age (<5%)
-   Sex (<5%)
-   Systolic blood pressure (25%)
-   Glasgow Coma Scale (25%)
-   Injury Severity Score (10%)
-   Mechanism of injury (10%)

These are known individual-level prognostic factors for the primary outcome. These covariates will be included in the models specified in @eq-logit-model and @eq-identity-model as fixed effects. We will model the continuous covariates assuming linear effects. We will assume no interaction between subgroups and secular time trends, which may be evaluated in future analyses if indicated.

[^additional-covariates]: **Questions:** 1. Should we adjust for calendar time because the last batch will be two years later than the first batch? 2. Should we add it as covariate in the list above? 3. Should it be continuous (e.g., months since the study started) or categorical calendar time (if you expect non-linear trends or stepwise changes over years)? 4. Should we analyse Time Since Admission? Depending on the nature of the trauma, the time since patient admission might also influence outcomes, especially if the timing of interventions or care decisions plays a critical role. Eg. time from admission to surgery.

### Subgroup analyses

We will perform the following subgroup analyses:

<!--[^too-many-subgroups]: **Question:** Are we exploring too many subgroups?-->

-   geographical region, defined using the state in which the participating hospital is located. Demonstrating the consistency of any effect across multiple regions will enhance the generalisibility of the results;
-   age groups, defined as older adolescents (15-19 years), young adults (20-24 years), adults (25-59 years), and older adults (60 years and older) \[@Diaz2021;
-   sex, using the levels male and female;
-   clinical cohorts, defined as blunt multisystem trauma, penetrating trauma, and severe isolated traumatic brain injury, with modification to avoid overlap between the cohorts; and
-   cluster size (number of patients per hospital)

These subgroup analyses will be conducted by adding the subgroup variable and the interaction between the subgroup variable and the intervention exposure variable as fixed effects to the models specified in @eq-logit-model and @eq-identity-model.

### Treatment of missing data

We will present the frequency and percentage of missing data for all variables. If the percentage of missing data for the primary outcome is less than 10%, we will perform a complete case analysis. If the percentage of missing data for the primary outcome is 10% or more, we will handle missing data depending on the missing data mechanism. If the data are missing at random (MAR), we will perform multiple imputation using multiple imputation by chained equations (MICE), imputing data for the primary outcome as well as all covariates included in the fully adjusted model. The number of imputations will be determined by the percentage of missing data, with a minimum of 20 imputations. If there is evidence that the data are missing not at random (MNAR), we will explore the impact of this assumption using a sensitivity analysis (e.g., pattern mixture models or selection models) to assess how robust our findings are to different assumptions about the missing data mechanism. Evidence that data are MNAR may include systematic differences in missingness across observed covariates, patterns suggestive of non-random dropout, or clinical expectations that missingness is related to unmeasured outcomes. Additionally, we will perform diagnostic checks after multiple imputation to ensure the quality of the imputation process, including comparing distributions of observed and imputed data and checking convergence.

<!-- 
Thoughts on missing data: 

Check nature of missing data, 
If MCAR, complete case analysis is generally acceptable, as there would be no systematic differences between those with and without missing data. 
If MAR, then MICE is appropriate.
If MNAR, MICE may not fully address the missingness. In such cases, a sensitivity analysis exploring MNAR models might be useful (e.g., pattern mixture models or selection models).

Perhaps to strengthen the missing data plan, we could consider adding a few lines:

If there is evidence that the data is missing not at random (MNAR), we will explore the impact of this assumption using a sensitivity analysis (e.g., pattern mixture models or selection models) to assess how robust our findings are to different assumptions about the missing data mechanism. Additionally, we will perform diagnostic checks after multiple imputation to ensure the quality of the imputation process, including comparing distributions of observed and imputed data and checking convergence.

----

Let us discuss this further.

-->

## Analysis of secondary outcomes

Unless otherwise stated, secondary outcomes will be analysed using the same mixed-effects modelling framework as specified for the primary outcome, including fixed effects for time and intervention exposure and random effects for cluster and cluster-by-period, with the link function and outcome distribution adapted to the scale of the outcome.

### All cause mortality within 24 hours, 30 days and three months of arrival at the emergency department

These outcomes will be analysed as dichotomous variables using the mixed-effects binomial model with the logit and identity links specified in @eq-primary-model, to estimate odds ratios and absolute risk differences.

### Quality of life within seven days of discharge, and at 30 days and three months of arrival at the emergency department, among survivors

Quality of life will be measured using the official and validated translations of the EQ5D5L. This tool assesses five dimensions of health-related quality of life: mobility, self-care, usual activities, pain/discomfort, and anxiety/depression. Each dimension is rated on a Likert scale from 1 to 5. There is also a visual analogue scale (VAS) ranging from 0 to 100.

For each of the five ordinal dimensions, we will use a mixed-effects cumulative logit (proportional odds) model:

$$
\text{logit}\{\Pr(Y_{bkti} \leq j)\}
= \mu_j + \beta_{bt} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt}
$$
{#eq-ordinal-model}

Where $\mu_j$ is the intercept for the $j$th category.

The VAS will be analysed as a continuous outcome using a linear mixed-effects model:

$$
\text{VAS}_{bkti}
= \mu + \beta_{bt} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt} + \epsilon_{bkti},
\quad \epsilon_{bkti} \sim \mathcal{N}(0,\sigma^2)
$$
{#eq-linear-model}

### Disability within seven days of discharge, and at 30 days and three months of arrival at the emergency department, among survivors

We will measure disability using the WHO Disability Assessment Schedule 2.0 (WHODAS 2.0) [@ustun_measuring_2010]. This tool assesses six domains of functioning: cognition, mobility, self-care, getting along, life activities, and participation. Each domain is rated on a likert scale from 1 to 5, with 1 indicating no difficulties and 5 indicating extreme difficulties. We will analyse each domain separately using a mixed effects ordinal model as specified in @eq-ordinal-model. We will also calculate a WHODAS 2.0 summary score using the method referred to as the "complex scoring" method. This method involves summing the item scores within each of the six domains, then summing the scores of all domains, and finally transforming the total score to a 0-100 scale. We will analyse the summary score using a linear mixed effects model as specified in @eq-linear-model.

### Return to work at 30 days and three months after arrival at the emergency department, among survivors

Return to work will be analysed as a dichotomous outcome using the mixed-effects binomial model specified in @eq-primary-model with the logit link.

### Length of emergency department stay

Length of emergency department stay will be analysed as a time-to-event outcome, defined as time from arrival at the emergency department to discharge. We will treat death as a competing event and therefore use a mixed-effects competing risks model, specifically a Fine and Gray subdistribution hazards model. The model is specified in @eq-competing-risks-model.

$$
h^{(s)}_{bk}(t)
=
h^{(s)}_{0}(t)
\exp\{\theta X_{bkt} + \beta_{bt} + \alpha_{bk}\}
$$
{#eq-competing-risks-model}

Where:

- $h^{(s)}_{bk}(t)$: the subdistribution hazard for discharge at time $t$ for patients in cluster $k$ in batch $b$.
- $h^{(s)}_{0}(t)$: the baseline subdistribution hazard function.
- $\theta$: under the Fine–Gray model, represents the log subdistribution hazard ratio for discharge associated with ATLS® exposure.
- All other terms are defined as in the primary analysis model.

### Length of hospital stay

Length of hospital stay will also be analysed as a time-to-event outcome, defined as time from arrival at the emergency department to hospital discharge, using the model specified in @eq-competing-risks-model.

### Intensive care unit admission

Intensive care unit admission will be analysed as a dichotomous outcome using the mixed-effects binomial model specified in @eq-primary-model with the logit link.

### Length of intensive care unit stay

Length of intensive care unit stay will also be analysed as a time-to-event outcome, defined as time from admission to the intensive care unit to discharge from the intensive care unit, using the model specified in @eq-competing-risks-model.

### Adherence to ATLS^®^ principles during initial patient resuscitation

We will assess adherence to ATLS® principles using a 14-item checklist that covers the key steps of the ATLS® primary survey.  Adherence will be measured as the proportion of completed items on this checklist. This outcome will be analysed using a mixed-effects beta regression model with a logit link:

We will assess adherence to ATLS® principles using a 14-item checklist that covers the key steps of the ATLS® primary survey. Adherence will be measured as the proportion of completed items on this checklist. This outcome will be analysed using a mixed-effects beta regression model with a logit link:


$$
g\{\mathbb{E}(Y_{bkti})\}
= \mu + \beta_{bt} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt}
$$
{#eq-beta-model}

Where $\mathbb{E}(Y_{bkti})$ denotes the expected adherence proportion. The intervention effect $\theta$ represents the difference in expected adherence proportion on the logit (link) scale. Because the observed adherence proportion may take values of 0 or 1, values will be rescaled to lie in the open interval $(0,1)$ prior to fitting the beta regression model. A beta regression precision (dispersion) parameter will be estimated as part of the model.

# References
