---
title: Statistical Analysis Plan
subtitle: "Effects of Advanced Trauma Life Support^®^ Training Compared to Standard Care on Adult Trauma Patient Outcomes: A Stepped-Wedge Cluster Randomised Trial"
titlepage-footer: "{{< var version_type >}} {{< var version >}}, {{< var date >}}"
bibliography: ../shared-assets/bibliography.bib
csl: nature.csl
---

```{r setup}
#| echo: FALSE

noacsr::source_all_functions()
attach(global_variables())
```

{{< pagebreak >}}

# Administrative information

## Study identifiers

```{r}
#| echo: FALSE

protocol.variables <- yaml::read_yaml("../protocol/_variables.yml")
```

-   Protocol version `r protocol.variables$version` dated `r protocol.variables$date`

{{< include ../shared-assets/study-identifiers.qmd >}}

## Changelog

Once version 1.0.0 is finalised, this section will be updated with a changelog.

<!--

| Version | Date       | Details | 
|---------|------------|---------|

-->

## Contributors

| Name and ORCID                                                                  | Affiliation                              | Role                       |
|---------------------------------------------------------------------------------|------------------------------------------|----------------------------|
| Martin Gerdin Wärnberg (MGW) `r insert_orcid_link("0000-0001-6069-4794")`       | Karolinska Institutet                    | Principal Investigator     |
| Anna Olofsson (AO) `r insert_orcid_link("0000-0002-9460-108X")`                 | Karolinska Institutet                    | Statistician, TMG member   |
| Karla Hemming (KH) `r insert_orcid_link("0000-0002-2226-6550")`                 | University of Birmingham, Birmingham, UK | Statistician, TMG member   |
| James Martin (JM) `r insert_orcid_link("0000-0002-6949-4200")`                  | University of Birmingham, Birmingham, UK | Statistician               |
| Jessica Kasza (JK) `r insert_orcid_link("0000-0002-8940-0136")`                      | Monash University, Melbourne, Australia | Statistician               |
| Arpita Ghosh (AG) `r insert_orcid_link("0000-0003-0036-4376")`                      | The George Institute for Global Health, New Delhi, India | Statistician               |

```{=latex}
% Pandoc renders markdown tables as longtable, which increments the table counter
% even without a caption. This table should not count toward Table numbering.
\addtocounter{table}{-1}
```

{{< pagebreak >}}

# Trial synopsis

{{< include ../shared-assets/synopsis.qmd >}}

{{< pagebreak >}}

# Special considerations

{{< include ../shared-assets/special-considerations.qmd >}}

{{< pagebreak >}}

# Aim and objectives

## Aim

To compare the effects of ATLS^®^ training with standard care on outcomes in adult trauma patients.

## Objectives

- To estimate the effects of ATLS^®^ training on 30-day in-hospital mortality compared to standard care.
- To assess the robustness of the main analysis results to different model specifications through sensitivity analyses.
- To estimate the effects of ATLS^®^ training on secondary outcomes compared to standard care.
- To explore the effects of ATLS^®^ training on subgroups of patients compared to standard care.

# Statistical analysis

## Design

<!-- sap-revision-todo D3 (comments id=3 MGW, id=4 AO; Round Two): confirm design descriptors including allocation and repeated measures on clusters. -->
This is a batched stepped-wedge cluster randomised trial, including a total of `r clusters` clusters (see @fig-trial-design). Each cluster is a hospital and represents the unit for randomisation, even if one or more units are trained in the same hospital. There are two intervention conditions: ATLS® training and standard care. The trial is designed and will be analysed within a superiority framework, with ATLS® training compared with standard care for the primary and secondary outcomes. 

The trial will be composed of `r batches` batches of identical `r total.months - 1`-period `r sequences`-sequence designs, with one cluster allocated to each sequence of each batch (equal allocation across sequences) [@Kasza2022]. Each period is one month, and each cluster will be in the trial for a total of `r total.months` months. Clusters are repeatedly measured across periods. The intervention will be implemented during a one-month transition period, which will be excluded from the analysis. There will be an anticipated overlap of `r batches.overlap.months` months between successive batches.

Clusters will be sequentially allocated to batches based on when they enter the study, and within each batch clusters will then be randomised to an intervention implementation sequence. Individual patients will be included cross-sectionally for the primary outcome, although some secondary outcomes will be assessed at multiple follow-up time points. The anticipated trial period is from February 2025 to October 2029.

```{r}
#| echo: false
#| label: fig-trial-design
#| fig.cap: "Trial design. Lines represent the duration of patient enrolment across clusters and phases. Clusters will be sequentially allocated to a batch based on when they enter the study. Within each batch clusters will then be randomised to an intervention implementation sequence."
#| fig.align: "center"
#| message: FALSE
#| warning: FALSE

knitr::include_graphics(
  create_trial_design_flowchart(
    clusters = clusters, sequences = sequences, batches = batches,
    min.standard.care.months = min.standard.care.months,
    min.intervention.months = min.intervention.months,
    batches.overlap.months = batches.overlap.months,
    transition.months = transition.months,
    transition.overlap.months = transition.overlap.months,
    current.month = NULL,
    return.figure = FALSE,
    device = "png"
  )
)
```

Because certain secondary outcomes will be more burdensome to collect and because we anticipate larger effects on these outcomes, we will nest a staircase design within the main stepped-wedge design to measure these outcomes (see @fig-nested-staircase-design) [@grantham_staircase_2024]. The staircase design will include a random subset of patients presenting during the three months preceding the transition phase, and the three months following the transition phase.  

```{r}
#| echo: false
#| label: fig-nested-staircase-design
#| fig.cap: "Nested staircase design. The design includes a random subset of patients presenting during the three months preceding the transition phase, and the three months following the transition phase."
#| fig.align: "center"
#| message: FALSE
#| warning: FALSE

knitr::include_graphics(
  create_trial_design_flowchart(
    clusters = clusters, sequences = sequences, batches = batches,
    min.standard.care.months = min.standard.care.months,
    min.intervention.months = min.intervention.months,
    batches.overlap.months = batches.overlap.months,
    transition.months = transition.months,
    transition.overlap.months = transition.overlap.months,
    staircase.months = staircase.months,
    return.figure = FALSE,
    device = "png"
  )
)
```

## Randomisation of clusters

We will enrol and assign clusters to batches once eligibility is confirmed and ethical approval is obtained. Each batch will include hospitals from different regions to optimise trial logistics and training delivery. Within each batch, clusters will be randomised to the intervention implementation sequences allocated to that batch. 

Randomisation is conducted at the cluster level and stratified by batch. Within each batch, allocation is balanced on cluster size, defined as the monthly volume of eligible patients, using covariate-constrained randomisation (CCR). CCR was selected to improve balance between intervention conditions on this key design variable.

For example, in Batch 1, five clusters were randomised to one of five sequences using CCR. A statistician (JM) implemented a computer algorithm in R to evaluate all possible allocations (n = 120) and calculate a balance metric reflecting how evenly cluster size would be distributed between intervention and control conditions. For each allocation, the squared difference in anticipated cluster size between conditions was calculated; because only one covariate was used, this squared difference constituted the balance statistic.

The 10% of allocations demonstrating the greatest imbalance were excluded, and the remaining allocations formed the candidate set. The final allocation was then selected at random from this set. Owing to limitations in the algorithm, the balance score was calculated assuming a complete design, although the trial design includes transition periods. This CCR procedure is repeated independently for each batch.

The random allocation sequence is generated by JM and communicated to the trial team to allow logistical arrangements, including securing places on upcoming ATLS^®^ training courses, to be initiated. Allocation order will be concealed from participating sites for as long as logistically feasible, recognising that advance planning is required to release physicians for training.

## Outcomes

A summary of all outcomes, analysis sets, effect measures, and handling of intercurrent events is provided in @tbl-outcomes-summary. 

<!-- sap-revision-todo C1 (comments id=55, id=56 KH; id=57, id=58 AO/MGW; Round Two id=11): death/transfer ICE per outcome; primary includes deaths after transfer. -->
{{< include ../shared-assets/outcomes.qmd >}}

```{r}
#| label: outcomes-table-data
#| include: false

outcomes.table <- jsonlite::fromJSON("tables/outcomes-summary.json")

names(outcomes.table) <- c(
  "Outcome",
  "Design component",
  "Data type",
  "Analysis set",
  "Main analysis model",
  "Effect measure",
  "Potential intercurrent events",
  "Strategy for handling intercurrent events"
)
```

```{r}
#| label: tbl-outcomes-summary
#| tbl-cap: "Summary of outcomes, analysis sets, effect measures, and handling of intercurrent events"
#| echo: false
#| message: false
#| warning: false
#| results: asis
ncol.outcomes <- ncol(outcomes.table)
# Column widths sum to 0.84\\linewidth so tabcolsep does not overflow the text block
col.widths <- c(0.20, 0.08, 0.07, 0.09, 0.13, 0.08, 0.10, 0.09)
col.align <- sprintf("p{%s\\linewidth}", col.widths)

outcomes.kable <- knitr::kable(
  outcomes.table,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
  align = col.align
) |>
  kableExtra::kable_styling(
    latex_options = c("repeat_header"),
    font_size = 8
  )

# kableExtra repeat_header uses the wrong colspan for p{} column alignments
outcomes.latex <- as.character(outcomes.kable)
outcomes.latex <- gsub(
  sprintf("\\\\multicolumn\\{%d\\}", ncol.outcomes * 2 - 2),
  sprintf("\\\\multicolumn{%d}", ncol.outcomes),
  outcomes.latex
)

# Margins for the landscape table pages. Because pdflscape rotates the page,
# geometry's top/bottom control the VISUAL left/right (the long edge) and
# left/right control the VISUAL top/bottom. \newgeometry must come BEFORE
# \begin{landscape} so the landscape text block is built from these margins.
cat("\\newgeometry{top=2cm,bottom=2cm,left=2cm,right=2cm}\n")
cat("\\begin{landscape}\n")
cat("\\setlength{\\tabcolsep}{3pt}\n")
cat("\\setlength{\\LTleft}{\\fill}\n")
cat("\\setlength{\\LTright}{\\fill}\n")
cat(outcomes.latex)
cat("\\end{landscape}\n")
cat("\\restoregeometry\n")
```

## Statistical hypotheses

The primary analysis will estimate the effect of ATLS training on 30-day in-hospital mortality compared with standard care. The treatment effect will be expressed as an odds ratio (OR) and as an absolute risk difference (ARD). An OR < 1 or ARD < 0 indicates lower mortality among patients treated in hospitals after ATLS implementation compared with standard care. Based on our pilot study and systematic review of the literature, we expect the intervention to reduce mortality.

## Sample size calculations

<!-- sap-revision-todo D4 (comments id=73 KH, id=74 AO, id=75 MGW): note design-phase Shiny CRT Calculator; no protocol cross-ref. -->
Sample size calculations were performed during the trial design phase using the Shiny CRT Calculator [@Hemming2020Feb].

### Main stepped wedge design and primary outcome

{{< include ../shared-assets/sample-size-calculations-main.qmd >}}

### Nested staircase design and secondary outcomes

{{< include ../shared-assets/sample-size-calculations-nested.qmd >}}

## Statistical principles

### Statistical software

We will use the R Statistical Software for all analyses [@R].

### Levels of statistical significance and confidence

We will use a two-sided significance level of 0.05 for all analyses, and we will report 95% confidence intervals (CI) for all estimates. We will not adjust for multiple testing because no secondary outcome is regarded as singularly more important.

<!-- sap-revision-todo G2 (comment id=5 MGW; Round Two): single final analysis once all data are available. -->
Primary and secondary outcomes will be analysed collectively in a single final analysis once all trial data are available.

## Analysis sets

<!-- sap-revision-todo A4 (comments id=102 KH, id=103 AO): clarify hospital→unit→patient hierarchy vs model RE; justify omitting unit-level RE; standardise on "units" (protocol term), not "teams". -->
<!-- sap-revision-todo H4 (comments id=12 MGW, id=13 AO; Round Two): specify target population per analysis including LOS/ICU and nested survivors. -->
The trial has a hierarchical structure in which hospitals represent the clusters, units within hospitals deliver care, and individual patients constitute the unit of observation and analysis. Randomisation is performed at the hospital level, with all trained units within the same hospital treated as part of the same cluster. Analysis models therefore account for clustering at the hospital level and do not include a separate random effect for unit within hospital. The primary analysis set will include all observations within clusters with available data on the primary outcome and irrespective of the receipt of the intervention. 

Clusters and observations within clusters will be considered exposed to the intervention after the date at which the cluster was scheduled to transition. All data will be included with the exception of the transition phases. Group allocation for a patient depends on the period in which the patient was admitted to the hospital. If clusters deviate considerably from the scheduled date of transition, defined as more than four weeks, we will conduct a sensitivity analysis using the actual date of transition to define exposure. 

The analysis set will be specified for each analysis. For the primary outcome, mortality secondary outcomes, length-of-stay outcomes, and ICU admission, the analysis set is all eligible patient participants included in the main stepped-wedge trial (with length of ICU stay restricted to patients admitted to ICU). For quality-of-life and disability outcomes, the analysis set is patients alive at the relevant follow-up time point within the nested staircase sample. For return-to-work outcomes, the analysis set is patients alive at the relevant follow-up time point in the main stepped-wedge trial for whom return to work is relevant. For adherence to ATLS® principles, the analysis set is the random sample of patients observed within the nested staircase design. 

The analysis set for each outcome will include all available participants from the relevant target population with observed outcome data, subject to any restrictions imposed by the outcome definition or the nested staircase sampling scheme. For adjusted analyses under the available-case approach, the analysis set will be restricted to participants with complete covariate data, as described in @sec-treatment-of-missing-data.

<!-- sap-revision-todo H3 (comment id=9 MGW; Round Two): stepped-wedge CONSORT extension for flow diagram. -->
We will use a CONSORT diagram to display the flow of hospitals, clusters and patients through the trial, see @fig-consort-diagram. We will present cluster level summaries of the intervention effect. We will report the study according to the CONSORT guidelines for stepped-wedge randomised trials [@Hemming2018].

```{r}
#| echo: false
#| label: fig-consort-diagram
#| fig.cap: "Consort diagram. The diagram displays the flow of clusters and participants through the trial."
#| fig.align: "center"
#| message: FALSE
#| warning: FALSE

knitr::include_graphics(
  create_consort_diagram(
    sequences = sequences,
    periods = total.months,
    return.figure = FALSE,
    device = "png"
  )
)
```

## Baseline analyses

### Cluster characteristics

We will describe cluster characteristics including location and size using frequencies and percentages for discrete variables and medians and interquartile ranges (Q1-Q3) for continuous variables. See @tbl-cluster-characteristics for details.

```{r}
#| label: tbl-cluster-characteristics
#| tbl-cap: "Cluster characteristics"
#| echo: false
#| message: false
#| warning: false
create_cluster_characteristics_table()
```

### Patient characteristics

<!-- sap-revision-todo E1 (comments id=166 KH, id=167 AO): all eligible patients over entire trial; baseline = characteristics not outcomes. -->
We will describe patient characteristics among all eligible patients over the entire trial period, stratified by before and after ATLS training and overall, using frequencies and percentages for discrete variables and medians and interquartile ranges (Q1-Q3) for continuous variables. Missing values will be reported for each characteristic. We will not adjust for clustering when presenting baseline characteristics (i.e. non-outcome variables). We will present key characteristics in the results section, and present all characteristics as supplementary tables. See @tbl-patient-characteristics for key characteristics and @tbl-patient-characteristics-all for all characteristics.

```{r}
#| label: tbl-patient-characteristics
#| tbl-cap: "Key patient characteristics"
#| echo: false
#| message: false
#| warning: false
create_patient_characteristics_table()
```

### Outcomes

Summaries of primary and secondary outcomes will be presented in the results section, stratified by before and after ATLS training and overall, using the same summary statistics as the patient characteristics tables. Missing values will be reported for each outcome. See @tbl-outcomes-descriptive for details.

```{r}
#| label: tbl-outcomes-descriptive
#| tbl-cap: "Descriptive summaries of outcomes"
#| echo: false
#| message: false
#| warning: false
#| results: asis
create_outcomes_descriptive_table()
```

## Interim analysis

{{< include ../shared-assets/interim-analysis.qmd >}}

<!-- 
Removed the recalculation of the sample size from the interim analysis after discussion with Karla and James on 2024-10-16. The rationale was that the estimates would be associated with so much uncertainty and that it wouldn't change things.

We will recalculate the sample size based on the observed period sizes and the observed effect size, using the Shiny CRT Calculator [@Hemming2020Feb]. In the revised sample size calculation, we will assume an ICC of 0.02, but consider sensitivity across the range 0.01-0.05 [@Campbell2005; @Eldridge2015], and a CAC of 0.9, but consider sensitivity across the range 0.8-1.0 [@Hemming2020Feb; @Martin2016; @Korevaar2021]. These assumptions are the same as in the original sample size calculation. -->

## Analysis of the primary outcome

The primary outcome is in-hospital mortality within 30 days of arrival at the emergency department and will be analysed as a dichotomous variable. We will estimate the intervention effect as the odds ratio (OR) of death between the ATLS® and standard care arms, with an OR < 1 indicating lower odds of death in the ATLS® arm compared to the standard care arm and vice versa.  We will also estimate the absolute risk difference (ARD) of death between the ATLS® and standard care arms, with an ARD < 0 indicating lower risk of death in the ATLS® arm compared to the standard care arm and vice versa.

<!-- sap-revision-todo B8 (comment id=20 AO; Round Two 27a): confirm hospital-level clustering, outcome-appropriate links, and cluster-specific intervention effect. -->
We will use a mixed-effects binomial regression model to estimate the effect of ATLS® on mortality. The OR will be estimated using the logit link, and the ARD will be estimated using the identity link. These models account for clustering at the hospital level and provide cluster-specific estimates of the intervention effect. The primary analysis will be based on the most complex model that converges from a prespecified sequence of models of decreasing complexity. The analysis will proceed from the most complex model to progressively simpler models if there are convergence or numerical stability issues. The primary analysis will then be subject to sensitivity analyses. If the binomial model with the logit link converges but the corresponding model with the identity link does not, then we will report only an OR. 

All models in this sequence will include a fixed effect for intervention exposure and adjust for time, although the parametrisation of time and the random effects structure may be simplified.

The model sequence is defined as follows:

1. **Primary model**: a model including fixed effects for period, modelled as a categorical variable with separate period effects for each batch, and a fixed effect for intervention exposure. Clustering will be accounted for by including a random intercept for cluster and a random cluster-by-period effect, both assumed to follow a normal distribution. The model is specified in @eq-primary-model.

$$
g\{\Pr(Y_{bkti} = 1)\} = μ + β_bt + θ X_bkt + α_bk + γ_bkt
$$ {#eq-primary-model}

2. **Simplified random effects structure**: a model including fixed effects for period and intervention exposure and a random intercept for cluster only, thereby removing the random cluster-by-period effect. The model is specified in @eq-simplified-random-effects-model.

$$
g\{\Pr(Y_{bkti} = 1)\} = μ + β_bt + θ X_bkt + α_bk
$$ {#eq-simplified-random-effects-model}

<!-- sap-revision-todo B1 (comments id=191 KH, id=192 AO): state retained random effects and batch indexing for the shared period effects model. -->
3. **Model with shared period effects**: a model in which fixed period effects are shared across batches, with time adjusted for using a common period effect or calendar time modelled directly (e.g., as a continuous covariate or using splines). A random intercept for cluster and a random cluster-by-period effect are retained; clusters remain indexed by batch, but the fixed period effects are common across batches. The model is specified in @eq-shared-period-model.

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \beta_{t} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt}
$$ {#eq-shared-period-model}

4. **Combined simplified model**: a model that includes both a simplified random effects structure (random intercept for cluster only) and shared period effects. This represents the most parsimonious mixed-effects model considered acceptable for the primary analysis. The model is specified in @eq-combined-simplified-model.

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \beta_{t} + \theta X_{bkt} + \alpha_{bk}
$$ {#eq-combined-simplified-model}

Where:

- $g(\cdot)$: the link function, specified as either the logit link, $g(p) = \log\{p/(1-p)\}$, or the identity link, $g(p) = p$, depending on the estimand of interest.
- $\theta$: **the intervention effect**, represented as a fixed effect of intervention exposure (the primary estimand); under the logit link this represents the log-odds ratio for death associated with ATLS® exposure compared with standard care, and under the identity link it represents the absolute risk difference.
- $\Pr(Y_{bkti} = 1)$: the probability of death for patient $i = 1, \dotsc, m$ in cluster $k = 1, \dotsc, `r clusters`$ during period $t = 1, \dotsc, `r total.months - 1`$ in batch $b = 1, \dotsc, `r batches`$.
- $\mu$: the intercept, representing the value of the linear predictor for the reference levels of all fixed effects, with random effects equal to zero.
- $\beta_t$: the fixed effect of period $t$, shared across batches.
- $\beta_{bt}$: the fixed effect of period $t$ in batch $b$, accounting for a separate period effect for each batch, resulting in a total of `r batches` $\times$ (`r total.months - 1`) period effects.
- $X_{bkt}$: the intervention exposure indicator for cluster $k$ during period $t$ in batch $b$, with $X_{bkt} = 1$ for ATLS® and $X_{bkt} = 0$ for standard care.
- $\alpha_{bk}$: a normally distributed random effect for cluster $k$ in batch $b$, representing between-cluster (hospital-level) variability in the outcome.
- $\gamma_{bkt}$: a normally distributed random effect for cluster $k$ in period $t$ in batch $b$, representing within-cluster variation over time and capturing cluster-specific deviations from the average period effect.

The distributional assumptions for the random effects are specified in @eq-distributional-assumptions.

$$
\begin{aligned}
\alpha_{bk} &\overset{\text{iid}}{\sim} \mathcal{N}(0,\sigma^2_{\alpha}), \\
\gamma_{bkt} &\overset{\text{iid}}{\sim} \mathcal{N}(0,\sigma^2_{\gamma}), \\
\alpha_{bk} &\perp \gamma_{bkt}.
\end{aligned}
$$ {#eq-distributional-assumptions}

To correct potential inflation of the type I error rate due to the small number of clusters, a small-sample correction will be applied. Our primary method will be the Kenward–Roger correction [@grantham_evaluating_2022], but this choice may be updated based on methodological developments available closer to the end of patient inclusion. The correction applied may also differ for outcomes collected via the complete stepped-wedge design and the incomplete nested staircase designs.

<!-- sap-revision-todo B3 (comments id=21, id=22 AO; Round Two 27c/27d): model-based SEs, small-sample adjustments, Wald-type CIs. -->
Standard errors will be model-based. Small-sample adjustments, including degrees-of-freedom corrections such as Kenward–Roger, will be applied where relevant. Confidence intervals will be Wald-type.

<!-- sap-revision-todo A1 (comments id=199 KH, id=200 AO); B4 (comment id=23 AO; Round Two 27b): general estimation framework is ML with Laplace; package/implementation not prespecified. -->
The general estimation framework for the mixed-effects models is maximum likelihood, with a Laplace approximation used to integrate out the random effects. The specific R package and numerical implementation will be chosen based on convergence behaviour and the availability of appropriate small-sample corrections closer to the time of analysis. If the binomial model with the logit link converges but the corresponding model with the identity link does not, then we will report only an OR. 

<!-- sap-revision-todo B2 (comments id=24 MGW, id=25 AO; Round Two): state that diagnostics will check distributional assumptions and convergence. -->
Model diagnostics will be used to assess distributional assumptions and model convergence.

<!-- sap-revision-todo B5 (comments id=26, id=27 AO; Round Two): report time-adjusted within-cluster correlations, variance components, and latent-scale correlations (protocol-aligned). -->
We will report time-adjusted within-cluster correlations for all outcomes with 95% CIs, including the intra-cluster correlation (ICC) and within- and between-period correlations implied by the assumed correlation structures, together with all estimated variance components. For binary outcomes, correlations will also be reported on the latent scale.

<!-- sap-revision-todo B6 (comments id=28, id=29 AO; Round Two): ICC and related correlations derived from variance components; AR(1) rho from the AR(1) structure. -->
Intra-cluster and related correlation parameters will be derived from the estimated variance components of the fitted models. For models with an AR(1) within-cluster structure, the correlation parameter will be estimated directly from the AR(1) structure.

<!--
*Question:** Should we adjust for calendar time because the last batch will be two years later than the first batch?

JK: Would this be instead of the beta_bt terms? That is, instead of separate period effects for each batch, just terms for each period would be included? (This is related to the question about 78 parameters in the model). The separate period effects for each batch will adjust for the calendar time. This is the model that makes the fewest assumptions about the period effects. 

We've done a bit of work recently which shows that if there are separate period effects in each batch but instead shared period effects are included, this can lead to quite a lot of bias in the estimation of the treatment effect. This work also seems to indicate that when separate period effects are included but the period effects are actually shared, this doesn't really have any deleterious impacts on the inference for the treatment effect. (However note that parameterisation of time wasn't the focus of the work we've done).

So my recommendation would be for the primary analysis to be as currently specified (separate period effects for each batch); then if there are issues with fitting that model, try a model with calendar time (i.e. period effects shared when there is overlap between successive batches).

**Question:** Just a thought: keep in mind the complexity of the random effects structure; if it leads to convergence problems or overfitting, should we consider simplifying the random effects?

-->

### Sensitivity analyses

We will conduct sensitivity analyses to assess the robustness of the primary analysis results to different modelling assumptions. In all sensitivity analyses, the focus will be on the stability of the estimated intervention effect $\theta$. Unless otherwise stated, each sensitivity analysis will be operationalised using two separate models, one with the logit link and one with the identity link.

#### Models with autoregressive within-cluster correlation

We will explore whether models with more complex correlation structures provide a better fit to the data. These models are not used as primary analysis models because there is limited understanding of when such models converge and how best to choose between different plausible correlation structures. Instead, we include these models as sensitivity analyses.

We will first explore whether allowing temporal correlation of outcomes within clusters affects inference on the intervention effect. Specifically, we will consider a discrete time-decay correlation structure by specifying a random cluster effect with an autoregressive structure of order one (AR(1)), described in @eq-ar1. This parameterisation assumes that correlations between cluster-level random effects decay geometrically with increasing temporal separation between periods.

$$
\alpha_{bk,t} = \rho \alpha_{bk,t-1} + \epsilon_{t}, \quad \epsilon_{t} \sim \mathcal{N}(0, \sigma^2_\alpha)
$$ {#eq-ar1}

Where:

- $\rho$ is the correlation parameter governing the association between adjacent periods within the same cluster.
- $\alpha_{bk,t}$ is the random effect for cluster $k$ in period $t$ in batch $b$.
- $\epsilon_t$ is a normally distributed error term.

We note that this AR(1) formulation differs slightly from alternative parameterisations used elsewhere in which correlations are specified through a joint covariance matrix. Although both approaches capture decaying temporal correlation, the parameter $\rho$ is not directly equivalent to parameters used in other specifications and should not be interpreted interchangeably.

<!-- sap-revision-todo A6 (comments id=212 KH, id=213 AO): removed "Models with batch-specific secular trends" sensitivity; primary model already uses fixed batch-specific period effects β_bt (protocol-aligned); shared period effects remain in the convergence sequence. -->

#### Models with random cluster by intervention effects

<!-- sap-revision-todo A3 (comments id=220 KH, id=221 AO): dropped the over-complex three-way correlated RE section; retained a simplified random intervention slope with non-zero Cov(α,u) only (γ independent) to adhere to the protocol while addressing reviewer complexity concerns. -->
To assess sensitivity to the assumption of a common intervention effect across clusters, we will extend the primary analysis model to allow the intervention effect to vary by cluster. Specifically, we will include a random cluster-level slope for the intervention exposure, with a non-zero covariance between this slope and the cluster random intercept (as anticipated in the protocol). The cluster-by-period random effect remains independent of both. The model is specified in @eq-model-random-cluster-intervention.

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \beta_{bt} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt} + u_{bk} X_{bkt}
$$ {#eq-model-random-cluster-intervention}

Where:

- $u_{bk}$ is a random effect representing cluster-specific deviations from the average intervention effect.
- The cluster intercept and intervention slope are jointly normally distributed,
$$
\begin{pmatrix}
\alpha_{bk} \\
u_{bk}
\end{pmatrix}
\sim
\mathcal{N}\!\left(
\begin{pmatrix}
0 \\
0
\end{pmatrix},
\begin{pmatrix}
\sigma^2_{\alpha} & \sigma_{\alpha u} \\
\sigma_{\alpha u} & \sigma^2_{u}
\end{pmatrix}
\right),
$$
with $\sigma_{\alpha u} \neq 0$ allowed.
- $\gamma_{bkt} \sim \mathcal{N}(0,\sigma^2_{\gamma})$ is independent of $(\alpha_{bk}, u_{bk})$.

#### Models with time modelled with a spline function

We will further explore the potential for a time-varying treatment effect [@kenny_analysis_2022]. To explore if the fixed period effect is both parsimonious and adequate to represent the extent of any underlying secular trend, we will model calendar time using a shared natural cubic spline basis across all batches with knots at the equally spaced time points 3, 6 and 9. This will result in five spline basis functions, because the natural cubic splines are modelled with three degrees of freedom but are constrained to be linear before the first and after the last knot. The model is specified in @eq-spline-model.

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \sum_{j=1}^{5} \beta_j S_j(t, \{3,6,9\}) + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt}
$$ {#eq-spline-model}

Where:

- $t$ denotes calendar time (period index) common to all batches.
- $S_j(t, \{3,6,9\})$ denotes the $j$th natural cubic spline basis function, shared across batches.
- $\beta_j$ is the coefficient for the $j$th spline basis function.

#### Models exploring lag and weaning effects

We will also extend the primary analysis model to include an interaction between treatment and number of periods since first treated, to examine if there is any indication of a relationship between duration of exposure to the intervention and outcomes. This will allow us to model different lag effects (whereby it takes time for the intervention to become embedded within the culture before its impact can properly start to be realised); as well as weaning effects (whereby the effect of the intervention starts to decrease – or fade). This type of analysis attempts to disentangle how some clusters end up having a long exposure to the intervention and others have a much shorter exposure time. The model is specified in @eq-lag-weaning-model.

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \beta_{bt} + \theta X_{bkt}
+ \theta_{\text{int}} X_{bkt} \times T_{bkt}
+ \alpha_{bk} + \gamma_{bkt}
$$ {#eq-lag-weaning-model}

Where $\theta_{\text{int}}$ represents the interaction between intervention exposure and time since first treated, and $T_{bkt}$ is the number of periods since first exposure.

### Adjusted analyses

Fully adjusted covariate analysis will additionally adjust for<!--[^additional-covariates]--> (expected % missing data in parentheses):

-   Age (<5%)
-   Sex (<5%)
-   Systolic blood pressure (25%)
-   Glasgow Coma Scale (25%)
-   Injury Severity Score (10%)
-   Mechanism of injury (10%)

These are known individual-level prognostic factors for the primary outcome. These covariates will be included in the models specified in @eq-primary-model as fixed effects. We will model the continuous covariates assuming linear effects. We will assume no interaction between subgroups and secular time trends, which may be evaluated in future analyses if indicated.

<!--
[^additional-covariates]: **Questions:** 1. Should we adjust for calendar time because the last batch will be two years later than the first batch? 2. Should we add it as covariate in the list above? 3. Should it be continuous (e.g., months since the study started) or categorical calendar time (if you expect non-linear trends or stepwise changes over years)? 4. Should we analyse Time Since Admission? Depending on the nature of the trauma, the time since patient admission might also influence outcomes, especially if the timing of interventions or care decisions plays a critical role. Eg. time from admission to surgery.
-->

### Subgroup analyses

<!-- sap-revision-todo F1 (comments id=33 MGW, id=34 AO; id=368 KH): subgroup analyses of primary outcome via interactions. -->
<!-- sap-revision-todo F2 (comments id=243, id=246 KH): clarify subgroup coding; cluster size categorical; keep six subgroups. -->
We will perform the following subgroup analyses of the primary outcome:

-  geographical region, defined using the state in which the participating hospital is located, with each state as a separate category;
-  age groups, defined as older adolescents (15-19 years), young adults (20-24 years), adults (25-59 years), and older adults (60 years and older) [@Diaz2021];
-  sex, using the levels male and female;
-  clinical cohorts, defined as blunt multisystem trauma, penetrating trauma, and severe isolated traumatic brain injury, with modification to avoid overlap between the cohorts [@hornor_quality_2018]; this subgroup analysis will include only patients classified into one of these three cohorts;
-  patients with major trauma, defined as patients with an injury severity score (ISS) of 16 or higher versus those with ISS below 16;
-  cluster size, defined as the monthly volume of eligible patients (as used for covariate-constrained randomisation) and categorised as small (fewer than 12 patients per month), medium (12–20), or large (more than 20)

These subgroup analyses will be conducted by adding the subgroup variable and the interaction between the subgroup variable and the intervention exposure variable as fixed effects to the models specified in @eq-primary-model.

### Treatment of missing data {#sec-treatment-of-missing-data}

<!-- sap-revision-todo A5 (comment id=260 AO): 10% primary-outcome missingness chooses available-case pipeline vs MICE; do not mix complete-case adjusted analyses with covariate imputation. -->
We will present the frequency and percentage of missing data for all variables. Handling of missing data for the primary analysis will depend on the percentage of missing data for the primary outcome.

If less than 10% of participants have a missing primary outcome, we will use an available-case approach. The primary unadjusted analysis will include all participants with observed primary outcome data. Adjusted analyses will be restricted to participants with complete data on the adjustment covariates, and the unadjusted analysis will be repeated in this complete-case population to allow the results of the unadjusted and ajusted analyses to be directly compared.

<!-- sap-revision-todo B7 (comments id=31 MGW, id=32 AO; Round Two): MI model reflects hospital clustering plus period and intervention exposure. -->
If 10% or more of participants have a missing primary outcome, we will use multiple imputation by chained equations (MICE) under a missing-at-random assumption, imputing the primary outcome, secondary outcomes as required for analysis, and the covariates included in the fully adjusted model, with a minimum of 20 imputations. The imputation model will reflect the hierarchical data structure, including clustering at the hospital level and design variables such as period and intervention exposure. Unadjusted and adjusted analyses will then be conducted using the imputed datasets. We will also present the corresponding complete-case analyses for comparison. Secondary outcomes will not be imputed under the available-case approach.

If there is evidence that data are missing not at random, we will explore this in a sensitivity analysis.

<!-- 
Thoughts on missing data: 

Check nature of missing data, 
If MCAR, complete case analysis is generally acceptable, as there would be no systematic differences between those with and without missing data. 
If MAR, then MICE is appropriate.
If MNAR, MICE may not fully address the missingness. In such cases, a sensitivity analysis exploring MNAR models might be useful (e.g., pattern mixture models or selection models).

Perhaps to strengthen the missing data plan, we could consider adding a few lines:

If there is evidence that the data is missing not at random (MNAR), we will explore the impact of this assumption using a sensitivity analysis (e.g., pattern mixture models or selection models) to assess how robust our findings are to different assumptions about the missing data mechanism. Additionally, we will perform diagnostic checks after multiple imputation to ensure the quality of the imputation process, including comparing distributions of observed and imputed data and checking convergence.

----

Let us discuss this further.

-->

## Analysis of secondary outcomes

<!-- sap-revision-todo B2 (comments id=24 MGW, id=25 AO; Round Two): diagnostics for secondary models, including proportional odds for ordinal outcomes. -->
<!-- sap-revision-todo H1 (comments id=14, id=15, id=17 MGW/AO; Round Two): effect measures by outcome type; no HR after A2; cluster-specific estimates. -->
Unless otherwise stated, secondary outcomes will be analysed using the same mixed-effects modelling framework as specified for the primary outcome, including fixed effects for time and intervention exposure and random effects for cluster and cluster-by-period, with the link function and outcome distribution adapted to the scale of the outcome. Effect measures are odds ratios and absolute risk differences for binary outcomes, rate ratios for count (length-of-stay) outcomes, cumulative odds ratios for ordinal outcomes, mean differences for continuous outcomes, and a logit-scale difference for the adherence proportion; there are no time-to-event analyses and no hazard ratios. These models provide cluster-specific estimates of the intervention effect for patient-level outcomes. Model diagnostics will be used to assess distributional assumptions, the proportional-odds assumption for cumulative logit models, and model convergence.

### All cause mortality within 24 hours, 30 days and three months of arrival at the emergency department

These outcomes will be analysed as dichotomous variables using the mixed-effects binomial model with the logit and identity links specified in @eq-primary-model, to estimate odds ratios and absolute risk differences.

### Quality of life within seven days of discharge, and at 30 days and three months of arrival at the emergency department, among survivors

<!-- sap-revision-todo C5 (comments id=288, id=289 KH; id=290, id=291 AO): nested staircase collection; period effects on global timeline. -->
<!-- sap-revision-todo C6 (comments id=295, id=296 KH/AO; id=303 KH): clarify EQ-5D-5L yields ordinal domains plus VAS. -->
<!-- sap-revision-todo C7 (Anna email): standardise instrument name to EQ-5D-5L in SAP text. -->
Quality of life will be measured within the nested staircase design using the official and validated translations of the EQ-5D-5L. The EQ-5D-5L yields two outcome types: five ordinal domain scores (mobility, self-care, usual activities, pain/discomfort, and anxiety/depression), each rated on a Likert scale from 1 to 5, and a visual analogue scale (VAS) ranging from 0 to 100. For these analyses, period effects are defined on the global study timeline (not separately within each batch), and not all clusters contribute data in every period.

For each of the five ordinal domains, we will use a mixed-effects cumulative logit (proportional odds) model:

$$
\text{logit}\{\Pr(Y_{bkti} \leq j)\}
= \mu_j + \beta_{t} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt}
$$ {#eq-ordinal-model}

Where $\mu_j$ is the intercept for the $j$th category.

The VAS will be analysed as a continuous outcome using a linear mixed-effects model:

$$
\text{VAS}_{bkti}
= \mu + \beta_{t} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt} + \epsilon_{bkti},
\quad \epsilon_{bkti} \sim \mathcal{N}(0,\sigma^2)
$$ {#eq-linear-model}

### Disability within seven days of discharge, and at 30 days and three months of arrival at the emergency department, among survivors

<!-- sap-revision-todo C5 (comments id=288, id=289 KH; id=290, id=291 AO): nested staircase collection; period effects on global timeline. -->
Disability will be measured within the nested staircase design using the WHO Disability Assessment Schedule 2.0 (WHODAS 2.0) [@ustun_measuring_2010]. This tool assesses six domains of functioning: cognition, mobility, self-care, getting along, life activities, and participation. Each domain is rated on a likert scale from 1 to 5, with 1 indicating no difficulties and 5 indicating extreme difficulties. As for quality of life, period effects are defined on the global study timeline, and not all clusters contribute data in every period. We will analyse each domain separately using a mixed effects ordinal model as specified in @eq-ordinal-model. We will also calculate a WHODAS 2.0 summary score using the method referred to as the "complex scoring" method. This method involves summing the item scores within each of the six domains, then summing the scores of all domains, and finally transforming the total score to a 0-100 scale. We will analyse the summary score using a linear mixed effects model as specified in @eq-linear-model.

### Return to work at 30 days and three months after arrival at the emergency department, among survivors

Return to work will be analysed as a dichotomous outcome using the mixed-effects binomial model specified in @eq-primary-model with the logit link.

<!-- sap-revision-todo A2 (comments id=349, id=350 KH; id=351 AO; id=348 AO): replace Fine–Gray competing-risks LOS models with mixed-effects negative binomial count models; clarify purpose, random/period effects, and death/transfer handling. -->
<!-- sap-revision-todo C2 (comment id=348 AO): ED stay defined as hours to ED exit; exit routes enumerated; all routes contribute observed duration. -->
### Length of emergency department stay

Length of emergency department stay will be analysed as a count outcome: the number of hours from arrival at the emergency department until ED exit. Exit may occur via ward admission, ICU admission, transfer to another hospital, death, or discharge home; all of these routes contribute the observed ED duration. The purpose of the analysis is to estimate the effect of ATLS® on mean ED length of stay. Death and transfer truncate the observed stay and are not treated as censoring; mortality is analysed separately as a primary and secondary outcome.

We will use a mixed-effects negative binomial regression model with a log link, including fixed effects for intervention exposure and batch-specific categorical period effects, a random intercept for cluster, and a random cluster-by-period effect, matching the primary analysis model structure where feasible. If this model does not converge, we will simplify the random-effects and period structure following the same sequence as for the primary outcome. The model is specified in @eq-length-of-stay-model. The intervention effect $\theta$ is the log rate ratio for ED length of stay associated with ATLS® exposure compared with standard care; we will also report the corresponding multiplicative effect on the mean count (rate ratio).

$$
\log\{\mathbb{E}(Y_{bkti})\}
= \mu + \beta_{bt} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt}
$$ {#eq-length-of-stay-model}

Where $Y_{bkti}$ is the length-of-stay count for patient $i$ in cluster $k$ during period $t$ in batch $b$, and all other terms are defined as in the primary analysis model. A dispersion parameter will be estimated as part of the negative binomial model.

### Length of hospital stay

Length of hospital stay will be analysed as a count outcome: the number of days from arrival at the emergency department to hospital discharge (or death in hospital). The analysis uses the mixed-effects negative binomial model in @eq-length-of-stay-model, with the same handling of death and transfer as for ED length of stay.

### Intensive care unit admission

<!-- sap-revision-todo C1 (comments id=55, id=56 KH; id=57, id=58 AO/MGW; Round Two id=11): state death/transfer handling for ICU admission. -->
Intensive care unit admission will be analysed as a dichotomous outcome using the mixed-effects binomial model specified in @eq-primary-model with the logit link. Patients who die or are transferred without ICU admission will be classified according to whether ICU admission occurred before that event.

### Length of intensive care unit stay

Length of intensive care unit stay will be analysed as a count outcome among patients admitted to ICU: the number of days from ICU admission to ICU discharge (or death in ICU). The analysis uses the mixed-effects negative binomial model in @eq-length-of-stay-model, with the same handling of death and transfer as for the other length-of-stay outcomes.

### Adherence to ATLS^®^ principles during initial patient resuscitation

<!-- sap-revision-todo C3 (comment id=361 KH): checklist assessed in both standard-care and ATLS® periods. -->
<!-- sap-revision-todo C4 (comments id=61 KH, id=62 AO, id=63 MGW): express adherence as percentage; 14/14 = 100%; duplicate paragraph already removed. -->
<!-- sap-revision-todo C5 (comments id=288, id=289 KH; id=290, id=291 AO): nested staircase collection; period effects on global timeline. -->
We will assess adherence to ATLS® principles within the nested staircase design, in both the standard care and ATLS® periods, using a 14-item checklist that covers the key steps of the ATLS® primary survey. Adherence will be measured as the percentage of completed checklist items (completion of all 14 steps corresponds to 100% adherence). For these analyses, period effects are defined on the global study timeline (not separately within each batch), and not all clusters contribute data in every period. This outcome will be analysed using a mixed-effects beta regression model with a logit link, with the percentage first expressed as a proportion on the 0–1 scale:

$$
g\{\mathbb{E}(Y_{bkti})\}
= \mu + \beta_{t} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt}
$$ {#eq-beta-model}

Where $\mathbb{E}(Y_{bkti})$ denotes the expected adherence proportion. The intervention effect $\theta$ represents the difference in expected adherence proportion on the logit (link) scale. Because the observed adherence proportion may take values of 0 or 1, values will be rescaled to lie in the open interval $(0,1)$ prior to fitting the beta regression model. A beta regression precision (dispersion) parameter will be estimated as part of the model.

# Supplementary tables

```{r}
#| label: tbl-patient-characteristics-all
#| tbl-cap: "All patient characteristics"
#| echo: false
#| message: false
#| warning: false
#| results: asis
create_patient_characteristics_table(all = TRUE)
```

# References {.numbered}
::: {#refs}
:::
