---
title: Statistical Analysis Plan
subtitle: "Effects of Advanced Trauma Life Support^®^ Training Compared to Standard Care on Adult Trauma Patient Outcomes: A Stepped-Wedge Cluster Randomised Trial"
titlepage-footer: "{{< var version_type >}} {{< var version >}}, {{< var date >}}"
bibliography: ../shared-assets/bibliography.bib
csl: nature.csl
---

```{r setup}
#| echo: FALSE

noacsr::source_all_functions()
attach(global_variables())
```

{{< pagebreak >}}

<!-- github-issue-39: section labels everywhere; @sec- rendered as Title (Section N) via filters/sec-title-xref.lua. -->

# Administrative information {#sec-administrative-information}
## Study identifiers {#sec-study-identifiers}
```{r}
#| echo: FALSE

protocol.variables <- yaml::read_yaml("../protocol/_variables.yml")
```

-   Protocol version `r protocol.variables$version` dated `r protocol.variables$date`

{{< include ../shared-assets/study-identifiers.qmd >}}

## Changelog {#sec-changelog}
<!-- github-issue-11: changelog table only after version 1.0.0. -->
We will add a changelog here once we finalise version 1.0.0.

## Contributors {#sec-contributors}
| Name and ORCID                                                                  | Affiliation                              | Role                       |
|---------------------------------------------------------------------------------|------------------------------------------|----------------------------|
| Martin Gerdin Wärnberg (MGW) `r insert_orcid_link("0000-0001-6069-4794")`       | Karolinska Institutet                    | Principal Investigator     |
| Anna Olofsson (AO) `r insert_orcid_link("0000-0002-9460-108X")`                 | Karolinska Institutet                    | Statistician, TMG member   |
| Karla Hemming (KH) `r insert_orcid_link("0000-0002-2226-6550")`                 | University of Birmingham, Birmingham, UK | Statistician, TMG member   |
| James Martin (JM) `r insert_orcid_link("0000-0002-6949-4200")`                  | University of Birmingham, Birmingham, UK | Statistician               |
| Jessica Kasza (JK) `r insert_orcid_link("0000-0002-8940-0136")`                      | Monash University, Melbourne, Australia | Statistician               |
| Arpita Ghosh (AG) `r insert_orcid_link("0000-0003-0036-4376")`                      | The George Institute for Global Health, New Delhi, India | Statistician               |

```{=latex}
% Pandoc renders markdown tables as longtable, which increments the table counter
% even without a caption. This table should not count toward Table numbering.
\addtocounter{table}{-1}
```

<!-- sap-revision-todo J8 (GitHub #5): generative AI disclosure in administrative information. -->
## Use of generative artificial intelligence {#sec-use-of-generative-artificial-intelligence}
We used generative artificial intelligence tools, including large language models accessed via Cursor, to help draft, revise, and format this statistical analysis plan. We reviewed and approved all substantive statistical content, and we take full responsibility for the final document.

{{< pagebreak >}}

# Introduction {#sec-introduction}
## Trial synopsis {#sec-trial-synopsis}
{{< include ../shared-assets/synopsis.qmd >}}

{{< pagebreak >}}

## Special considerations {#sec-special-considerations}
{{< include ../shared-assets/special-considerations.qmd >}}

{{< pagebreak >}}

# Aim and objectives {#sec-aim-and-objectives}
## Aim {#sec-aim}
To compare the effects of ATLS^®^ training with standard care on outcomes in adult trauma patients.

## Objectives {#sec-objectives}
- To estimate the effects of ATLS^®^ training on 30-day in-hospital mortality compared to standard care.
- To assess how robust the main analysis results are to different model specifications, exposure definitions, and missing-data assumptions, through sensitivity analyses.
- To estimate the effects of ATLS^®^ training on secondary outcomes compared to standard care.
- To explore the effects of ATLS^®^ training on subgroups of patients compared to standard care.

## Framework {#sec-framework}
<!-- github-issue-13: CRT SAP reporting paper — clarify superiority framework plus statistical target of inference. -->
We have designed and will analyse the trial within a superiority framework, with ATLS^®^ training compared with standard care for the primary and secondary outcomes. We target a cluster-specific (conditional) intervention effect among eligible patient participants, not a marginal (population-averaged) effect and not an equal-weight average across clusters.

<!-- github-issue-12: reorder sections to Round Two checklist sequence. -->

# Study design {#sec-study-design}
## Trial design {#sec-trial-design}
<!-- sap-revision-todo D3 (comments id=3 MGW, id=4 AO; Round Two): confirm design descriptors including allocation and repeated measures on clusters. -->
This is a batched stepped-wedge cluster randomised trial with `r clusters` clusters (see @fig-trial-design). Each cluster is one hospital, and we randomise hospitals, even when we train more than one unit of physicians in the same hospital. There are two intervention conditions: ATLS^®^ training and standard care.

The trial comprises `r batches` batches of identical `r total.months - 1`-period `r sequences`-sequence designs, with one cluster allocated to each sequence of each batch (equal allocation across sequences) [@Kasza2022]. Each period is one month, and each cluster will be in the trial for a total of `r total.months` months. We measure clusters repeatedly across periods. We will implement the intervention during a one-month transition period, which we exclude from the analysis. There will be an anticipated overlap of `r batches.overlap.months` months between successive batches.

Clusters will be sequentially allocated to batches based on when they enter the study, and within each batch clusters will then be randomised to an intervention implementation sequence. We include individual patients cross-sectionally for the primary outcome, although we assess some secondary outcomes at multiple follow-up time points. The anticipated trial period is from February 2025 to October 2029.

```{r}
#| echo: false
#| label: fig-trial-design
#| fig.cap: "Trial design. Each line shows how long we include patients in a cluster, by batch and sequence. Clusters will be sequentially allocated to a batch based on when they enter the study. Within each batch clusters will then be randomised to an intervention implementation sequence."
#| fig.align: "center"
#| message: FALSE
#| warning: FALSE

knitr::include_graphics(
  create_trial_design_flowchart(
    clusters = clusters, sequences = sequences, batches = batches,
    min.standard.care.months = min.standard.care.months,
    min.intervention.months = min.intervention.months,
    batches.overlap.months = batches.overlap.months,
    transition.months = transition.months,
    transition.overlap.months = transition.overlap.months,
    current.month = NULL,
    return.figure = FALSE,
    device = "png"
  )
)
```

Because certain secondary outcomes will be more burdensome to collect and because we anticipate larger effects on these outcomes, we will nest a staircase design within the main stepped-wedge design to measure these outcomes (see @fig-nested-staircase-design) [@grantham_staircase_2024]. The staircase design will include a random subset of patients presenting during the three months preceding the transition phase, and the three months following the transition phase.

```{r}
#| echo: false
#| label: fig-nested-staircase-design
#| fig.cap: "Nested staircase design. We include a random subset of patients presenting in the three months before and the three months after each cluster's transition phase."
#| fig.align: "center"
#| message: FALSE
#| warning: FALSE

knitr::include_graphics(
  create_trial_design_flowchart(
    clusters = clusters, sequences = sequences, batches = batches,
    min.standard.care.months = min.standard.care.months,
    min.intervention.months = min.intervention.months,
    batches.overlap.months = batches.overlap.months,
    transition.months = transition.months,
    transition.overlap.months = transition.overlap.months,
    staircase.months = staircase.months,
    return.figure = FALSE,
    device = "png"
  )
)
```

## Intervention and control treatment {#sec-intervention-and-control-treatment}
{{< include ../shared-assets/intervention-and-control-treatment.qmd >}}

## Randomisation and inclusion {#sec-randomisation-and-inclusion}
We will enrol and assign clusters to batches once eligibility is confirmed and ethical approval is obtained. Each batch will include hospitals from different regions to optimise trial logistics and training delivery. Within each batch, clusters will be randomised to the intervention implementation sequences allocated to that batch.

We randomise clusters, stratified by batch. Within each batch we balance allocation on cluster size, defined as the monthly volume of eligible patients, using covariate-constrained randomisation (CCR). We chose CCR to improve balance between intervention conditions on this key design variable.

For example, in Batch 1, five clusters were randomised to one of five sequences using CCR. JM wrote an R program to enumerate all possible allocations (n = 120) and calculate a balance metric reflecting how evenly cluster size would be distributed between the two intervention conditions. For each allocation JM calculated the squared difference in anticipated cluster size between conditions; because there was only one covariate, this squared difference was the balance statistic.

We excluded the 10% of allocations with the greatest imbalance and kept the rest as the candidate set, then drew the final allocation at random from that set. Because of limitations in the algorithm, we calculated the balance score assuming a complete design, although the trial design includes transition periods. We will repeat this CCR procedure independently for each remaining batch.

JM generates the random allocation sequence for each batch and sends it to the Trial Team so they can begin logistical arrangements, including securing places on upcoming ATLS^®^ training courses. We will conceal the allocation order from participating clusters for as long as is logistically feasible, because releasing physicians for training requires advance planning.

<!-- sap-revision-todo I4 (comments id=45 KH, id=46 AO): cross-ref nested staircase patient sampling. -->
We select patients for the nested staircase outcomes separately from cluster randomisation, by stratified random sampling by shift; see @sec-nested-staircase-design-and-secondary-outcomes.

## Sample size calculations {#sec-sample-size-calculations}
<!-- sap-revision-todo D4 (comments id=73 KH, id=74 AO, id=75 MGW): note design-phase Shiny CRT Calculator; no protocol cross-ref. -->
We calculated the sample size during the trial design phase using the Shiny CRT Calculator [@Hemming2020Feb].

### Main stepped-wedge design and primary outcome {#sec-main-stepped-wedge-design-and-primary-outcome}
{{< include ../shared-assets/sample-size-calculations-main.qmd >}}

### Nested staircase design and secondary outcomes {#sec-nested-staircase-design-and-secondary-outcomes}
{{< include ../shared-assets/sample-size-calculations-nested.qmd >}}

## Interim analysis {#sec-interim-analysis}
{{< include ../shared-assets/interim-analysis.qmd >}}

<!-- sap-revision-todo J4: content of the interim report to the joint SDMC (list only; detailed shells deferred). -->
The Trial Team will prepare the interim analysis report for the joint Trial Steering and Data Monitoring Committee (SDMC), and the Trial Management Group will approve it before circulation. The report will follow the standard SDMC report template in the charter, except that it reports outcome data as completeness only, and will include:

- trial progress: numbers of clusters included and dropping out; numbers of potentially eligible patients screened, patients included, patients who did not consent to out-of-hospital follow-up, and patients lost to follow-up, overall and by intervention condition;
- adherence to the randomisation schedule and reasons for non-adherence;
- summary statistics of included clusters and patient participants by intervention condition;
- completeness of primary and secondary outcome data, overall and by intervention condition;
- ATLS^®^ training delivery, including numbers of physicians trained in each hospital;
- protocol deviations and proposed protocol amendments;
- safety events, as defined for SDMC monitoring;
- relevant external evidence; and
- the SDMC recommendation.

We will not present or compare outcome values by intervention condition in the interim report, to avoid biasing the SDMC or Trial Management Group. The SDMC will use the interim report to advise whether the trial remains feasible and should continue as planned.

<!--
Removed the recalculation of the sample size from the interim analysis after discussion with Karla and James on 2024-10-16. The rationale was that the estimates would be associated with so much uncertainty and that it wouldn't change things.

We will recalculate the sample size based on the observed period sizes and the observed effect size, using the Shiny CRT Calculator [@Hemming2020Feb]. In the revised sample size calculation, we will assume an ICC of 0.02, but consider sensitivity across the range 0.01-0.05 [@Campbell2005; @Eldridge2015], and a CAC of 0.9, but consider sensitivity across the range 0.8-1.0 [@Hemming2020Feb; @Martin2016; @Korevaar2021]. These assumptions are the same as in the original sample size calculation. -->

## Final analysis {#sec-final-analysis}
<!-- sap-revision-todo G2 (comment id=5 MGW; Round Two): single final analysis once all data are available. -->
Primary and secondary outcomes will be analysed collectively in a single final analysis once all trial data are available.

# Statistical principles {#sec-statistical-principles}
## Statistical hypotheses {#sec-statistical-hypotheses}
We will estimate the effect of ATLS^®^ training on 30-day in-hospital mortality compared with standard care, as an odds ratio (OR) and as an absolute risk difference (ARD). An OR < 1 or ARD < 0 indicates lower mortality among patients treated in hospitals after ATLS^®^ implementation than under standard care. Based on our pilot study and systematic review of the literature, we expect the intervention to reduce mortality.

## Levels of statistical significance and confidence {#sec-levels-of-statistical-significance-and-confidence}
We will use a two-sided significance level of 0.05 for all analyses, and we will report 95% confidence intervals (CI) for all estimates.

## Type I error control {#sec-type-i-error-control}
We will not adjust for multiple testing because we do not regard any secondary outcome as singularly more important. We will interpret the secondary outcomes together and not over-emphasise a single statistically significant result when the others appear unchanged.

<!-- sap-coherence-review (R2-03, R1 conflict 5): descriptive-summary conventions were stated three to five times each and had begun to drift; hoisted here because they govern tables in Trial sample, Intercurrent events, Effect measures and Safety events. -->
## Descriptive summaries {#sec-descriptive-summaries}
In every descriptive table in this plan we summarise discrete variables as frequencies and percentages, and continuous and count variables as medians and interquartile ranges (Q1–Q3); where a variable fits neither, we use the summary statistic given in the patient characteristics tables. We report missing values for each variable, and we do not adjust any of these summaries for clustering. Where a table has a key and an all version, we present the key version in the results section and the all version as a supplementary table.

## Adherence and protocol deviations {#sec-adherence-and-protocol-deviations}
<!-- github-issue-21: draft adherence and protocol deviations from existing SAP/protocol text (checklist 17a/17b). -->

### Adherence to the intervention {#sec-adherence-to-the-intervention}
We define adherence to the intervention at the cluster and individual levels. It is distinct from the secondary outcome adherence to ATLS^®^ principles during resuscitation, which we define and analyse under @sec-outcomes and @sec-analysis-of-secondary-outcomes and do not use to exclude patients from the primary analysis.

At the cluster level, adherence means that ATLS^®^ training is delivered during the scheduled transition period for that cluster’s randomised sequence, and that the cluster does not implement ATLS^®^ training before its scheduled transition. We will assess cluster-level adherence using the training log kept by the Trial Team. We will summarise cluster-level adherence descriptively, including the number and proportion of clusters that completed training as scheduled and the reasons for any non-adherence to the randomisation schedule. We will also review adherence to the randomisation schedule at the interim analysis; see @sec-interim-analysis.

At the individual level, the cluster’s scheduled transition date defines exposure to the intervention, as specified under @sec-analysis-sets. We classify patients by their period of admission.

### Protocol deviations {#sec-protocol-deviations}
A protocol deviation is any departure from the approved trial protocol, ICH-GCP, or applicable regulations. Consistent with the protocol, we will report to the sponsor without delay, for assessment, any serious breach or deviation that significantly affects, or is likely to affect, participant safety or integrity or the reliability of trial data. We will document minor deviations — those that do not affect participant safety or integrity and do not materially affect the scientific value of the trial — in the trial records of the principal investigator and the sponsor.

We will summarise protocol deviations descriptively for the final analysis and review them at the interim analysis, overall and by intervention condition where informative, giving counts by type (cluster-level versus individual-level, and serious versus minor where classified) and narrative reasons. We will not exclude clusters or patients from the primary analysis because of a deviation; the primary analysis still uses scheduled intervention exposure and the outcome data available under the missing-data plan, as specified under @sec-analysis-sets.

# Trial sample {#sec-trial-sample}
## Eligibility criteria {#sec-eligibility-criteria}
Our trial has eligibility criteria at the cluster and patient participant levels. We summarise them below; the protocol gives full details.

### Cluster eligibility criteria {#sec-cluster-eligibility-criteria}
{{< include ../shared-assets/eligibility-criteria-cluster.qmd >}}

### Patient participant eligibility criteria {#sec-patient-participant-eligibility-criteria}
{{< include ../shared-assets/eligibility-criteria-patient-participant.qmd >}}

## Participant flow {#sec-participant-flow}
<!-- github-issue-22: two-figure CONSORT shells (cluster + patient), following batched SW reporting practice. -->
<!-- github-issue-20: nested-staircase patient flow reported separately from main CONSORT. -->

<!-- sap-revision-todo H3 (comment id=9 MGW; Round Two): stepped-wedge CONSORT extension for flow diagram. -->
We will use CONSORT diagrams to display the flow of clusters and patients through the main batched stepped-wedge trial, reported as two figures in line with recent batched stepped-wedge practice [@escp_eagle_2024]: a cluster-level diagram (@fig-consort-clusters) and a patient-level diagram (@fig-consort-patients). For each sequence, the diagrams will report numbers randomly assigned or included, losses and exclusions after randomisation (with reasons), and numbers analysed for the primary outcome. The patient diagram will also summarise flow before and after ATLS^®^ training. A separate figure (@fig-consort-nested-staircase) shows flow for the nested-staircase secondary outcomes (adherence to ATLS^®^ principles, EQ-5D-5L, and WHODAS 2.0): patients in the staircase periods, losses and exclusions, and numbers analysed, by sequence, before and after ATLS^®^ training, and overall. We will present cluster-level summaries of the intervention effect. We will report the study according to the CONSORT guidelines for stepped-wedge randomised trials [@Hemming2018].

## Baseline characteristics {#sec-baseline-characteristics}
### Cluster characteristics {#sec-cluster-characteristics}
<!-- sap-revision-todo J7: non-floating longtable so placement stays in document order. -->
<!-- github-issue-14: add state to cluster characteristics shell. -->
We will describe cluster characteristics including state and size, where the state categories depend on the final set of participating clusters (@tbl-cluster-characteristics).

### Patient characteristics {#sec-patient-characteristics}
<!-- sap-revision-todo E1 (comments id=166 KH, id=167 AO): all eligible patients over entire trial; baseline = characteristics not outcomes. -->
<!-- sap-revision-todo J7: non-floating longtable so placement stays in document order. -->
We will describe patient characteristics among all eligible patients over the entire trial period, stratified by before and after ATLS^®^ training and overall (@tbl-patient-characteristics for key characteristics, @tbl-patient-characteristics-all for all characteristics).

# Outcomes {#sec-outcomes}
@tbl-outcomes-summary summarises all outcomes, analysis sets, effect measures, and how we handle intercurrent events.

<!-- sap-revision-todo C1 (comments id=55, id=56 KH; id=57, id=58 AO/MGW; Round Two id=11): death/transfer ICE per outcome; primary includes deaths after transfer. -->
{{< include ../shared-assets/outcomes.qmd >}}

```{r}
#| label: outcomes-table-data
#| include: false

outcomes.table <- jsonlite::fromJSON("tables/outcomes-summary.json")

names(outcomes.table) <- c(
  "Outcome",
  "Design component",
  "Data type",
  "Analysis set",
  "Main analysis model",
  "Effect measure",
  "Potential intercurrent events",
  "Strategy for handling intercurrent events"
)
```

```{r}
#| label: tbl-outcomes-summary
#| tbl-cap: "Summary of outcomes, analysis sets, effect measures, and handling of intercurrent events"
#| echo: false
#| message: false
#| warning: false
#| results: asis
ncol.outcomes <- ncol(outcomes.table)
# Column widths sum to 0.84\\linewidth so tabcolsep does not overflow the text block
col.widths <- c(0.20, 0.08, 0.07, 0.09, 0.13, 0.08, 0.10, 0.09)
col.align <- sprintf("p{%s\\linewidth}", col.widths)

outcomes.kable <- knitr::kable(
  outcomes.table,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
  align = col.align
) |>
  kableExtra::kable_styling(
    latex_options = c("repeat_header"),
    font_size = 8
  )

# kableExtra repeat_header uses the wrong colspan for p{} column alignments
outcomes.latex <- as.character(outcomes.kable)
outcomes.latex <- gsub(
  sprintf("\\\\multicolumn\\{%d\\}", ncol.outcomes * 2 - 2),
  sprintf("\\\\multicolumn{%d}", ncol.outcomes),
  outcomes.latex
)

# Margins for the landscape table pages. Because pdflscape rotates the page,
# geometry's top/bottom control the VISUAL left/right (the long edge) and
# left/right control the VISUAL top/bottom. \newgeometry must come BEFORE
# \begin{landscape} so the landscape text block is built from these margins.
cat("\\newgeometry{top=2cm,bottom=2cm,left=2cm,right=2cm}\n")
cat("\\begin{landscape}\n")
cat("\\setlength{\\tabcolsep}{3pt}\n")
cat("\\setlength{\\LTleft}{\\fill}\n")
cat("\\setlength{\\LTright}{\\fill}\n")
cat(outcomes.latex)
cat("\\end{landscape}\n")
cat("\\restoregeometry\n")
```

# Intercurrent events, populations and summary measures {#sec-intercurrent-events-populations-and-summary-measures}
## Intercurrent events {#sec-intercurrent-events}
<!-- github-issue-23: draft ICE strategies from outcomes-summary table and existing outcome analysis text. -->
Intercurrent events occur after intervention exposure and affect whether an outcome occurs or how we interpret it. @tbl-outcomes-summary lists the anticipated events and the strategy we use for each outcome. For each outcome we will report counts and percentages of patients with the event before we assess that outcome, stratified by before and after ATLS^®^ training and overall. Denominators follow the analysis set for the outcome (main stepped-wedge trial or nested staircase sample). See @tbl-intercurrent-events.

At the cluster level we treat late or early transition relative to the scheduled date, incomplete delivery of ATLS^®^ training, and cluster dropout as protocol or adherence issues, as described under @sec-adherence-and-protocol-deviations, rather than as outcome-specific intercurrent events. We still define exposure for these clusters as specified under @sec-analysis-sets.

<!-- sap-revision-todo C1 (comments id=55, id=56 KH; id=57, id=58 AO/MGW; Round Two id=11): death/transfer handling for length of stay and ICU admission, stated once here. -->
At the individual level, patients cannot withdraw from the intervention. For length of emergency department, hospital, and intensive care unit (ICU) stay, the observed stay is the completed duration to exit; death and transfer truncate it, and we do not treat either as censoring. For ICU admission we record whether the patient was admitted to ICU before death or transfer; patients who die or are transferred without being admitted to ICU count as not admitted.

<!-- github-issue-23: transfer is part of the mortality outcome definitions, not an ICE. -->
- **Transfer to another hospital.** Relevant to length-of-stay and ICU outcomes. It is not an intercurrent event for mortality outcomes, because the primary outcome includes in-hospital deaths after transfer and all-cause mortality counts deaths regardless of care setting.
<!-- github-issue-23: death as ICE is death before that outcome's assessment. -->
- **Death.** Relevant to length-of-stay and ICU outcomes and to survivor-only outcomes. As an intercurrent event, death is death before assessment of the outcome in question; later death is not an intercurrent event for earlier outcomes. Quality of life, disability, and return to work are defined only among patients alive at the relevant follow-up (while-alive restriction). Those analyses therefore target effects in survivors at that time point; mortality is analysed separately as a primary and secondary outcome.
- **Withdrawal of consent for non-routine data collection.** Relevant only to outcomes that require opt-in follow-up: all-cause mortality after the initial admission (when not available from records), return to work, quality of life, and disability. It is also relevant to the primary outcome when death after transfer cannot be ascertained from records and requires contact with the patient, a representative, or the receiving hospital. Withdrawal is not listed for routinely recorded in-hospital outcomes (length of stay and ICU admission) or for adherence, which are collected under opt-out consent. Participants who withdraw contribute observed non-routine outcomes up to withdrawal unless they also request removal of already collected data; subsequent missing values are handled under the missing-data plan; see @sec-missing-data.
- **Loss to follow-up.** Relevant to the same contact-dependent outcomes as withdrawal. These events create missing outcome data and are handled under the missing-data plan.
- **Incomplete observation of resuscitation.** Relevant only to adherence. We score the checklist by observing the initial resuscitation up to one hour after the physician first sees the patient. Observation is incomplete when we do not observe that whole window — for example when the observer could not complete the 14-item checklist because care ended, the patient left the emergency department, or observation was interrupted. We exclude unobserved and incompletely observed resuscitations from the adherence analysis (available-case analysis of the observed nested sample), and we treat death during the observation window the same way.

Return to work is further restricted to patients for whom the outcome is relevant; that restriction defines the analysis set and is not an intercurrent event.

Where we use a while-alive strategy, we assume that, conditional on the analysis model and design factors, death before we assess the outcome does not induce selection bias beyond what analysing mortality separately already addresses; we will therefore interpret mortality and survivor outcomes together. We will weigh the assumptions behind each strategy when we interpret results and when we check model assumptions; see @sec-analysis-methods.

<!-- github-issue-23: ICE frequency shell grouped by outcome and applicable events. -->

## Population {#sec-population}
<!-- github-issue-24: draft target populations from analysis-set and eligibility text. -->
We define the treatment effects in this plan for the trial’s eligible clusters and eligible patient participants, as specified under @sec-eligibility-criteria.

- **Clusters.** We estimate effects for participating hospitals (clusters) that meet the cluster eligibility criteria and that we randomise within the batched stepped-wedge design. Units within a hospital are part of the same cluster; we do not define the estimands separately by unit.
- **Main stepped-wedge patient population.** For the primary outcome, all-cause mortality secondary outcomes, length of emergency department and hospital stay, and ICU admission, the target population is eligible trauma patients included in the main stepped-wedge trial during non-transition periods. For length of ICU stay the population is those patients who are admitted to ICU.
- **Survivor populations.** We target quality of life and disability effects among patients in the nested staircase sample who are alive at the relevant follow-up time point, and return-to-work effects among patients in the main stepped-wedge trial who are alive at the relevant follow-up and for whom return to work is relevant.
- **Nested adherence population.** We target the adherence-to-ATLS^®^-principles effect among the patients we randomly select for observation in the nested staircase design during the observation window.

@sec-analysis-sets specifies which patients from each target population enter a given analysis.

## Analysis sets {#sec-analysis-sets}
<!-- sap-revision-todo A4 (comments id=102 KH, id=103 AO): clarify hospital→unit→patient hierarchy vs model RE; justify omitting unit-level RE; standardise on "units" (protocol term), not "teams". -->
<!-- sap-revision-todo H4 (comments id=12 MGW, id=13 AO; Round Two): specify target population per analysis including LOS/ICU and nested survivors. -->
<!-- github-issue-24: keep operational analysis-set rules here; target populations moved above. -->
The trial is hierarchical: hospitals are the clusters, units within hospitals deliver care, and individual patients are the level of observation and analysis. We randomise hospitals and treat all trained units within a hospital as part of the same cluster. Our models therefore account for clustering at the hospital level and do not include a separate random effect for unit within hospital. The primary analysis set will include all observations within clusters that have primary outcome data, whether or not the patient received the intervention and whether or not a trained physician delivered their care.

We will treat clusters, and the observations within them, as exposed to the intervention after the cluster's scheduled transition date. We will include all data except the transition phases. A patient's intervention condition depends on the period in which they were admitted. We will also define exposure by the actual transition date in a sensitivity analysis; see @sec-exposure-defined-by-the-actual-transition-date.

We specify the analysis set for each analysis. It is the target population for that outcome under @sec-population, restricted to patients with observed outcome data and to any further outcome-specific constraints in @tbl-outcomes-summary. For adjusted analyses under the available-case approach we restrict the analysis set to patients with complete covariate data, as described in @sec-missing-data.

## Effect measures {#sec-effect-measures}
<!-- github-issue-25: revise effect measures against checklist; keep descriptive shells and cross-ref analysis shells. -->
<!-- sap-revision-todo J1: key vs all descriptive outcome shell tables (domain-level EQ-5D-5L/WHODAS in supplementary). -->
We measure all outcomes on individual patients, so these are patient-level summaries. We will summarise the primary and secondary outcomes stratified by before and after ATLS^®^ training and overall (@tbl-outcomes-descriptive for key outcomes, @tbl-outcomes-descriptive-all for all outcomes).

<!-- github-issue-13: CRT SAP reporting paper — participant-level vs cluster-average; cluster-specific vs marginal. -->
<!-- sap-revision-todo H1 (comments id=14, id=15, id=17 MGW/AO; Round Two): effect measures by outcome type; no HR after A2; cluster-specific estimates. -->
@tbl-outcomes-summary lists the comparative effect measures by outcome; we estimate them from the mixed-effects models in @sec-analysis-methods. These models give cluster-specific (conditional) estimates of the intervention effect for patient-level outcomes, not marginal (population-averaged) estimates. We target the patient-level average among eligible patients in participating clusters, rather than an equal-weight average of cluster-level summaries. The types of effect measures are as follows:

- **Dichotomous outcomes:** odds ratio (logit link) and absolute risk difference (identity link), where both links are used.
- **Count (length-of-stay) outcomes:** rate ratio from a negative binomial model with log link.
- **Ordinal outcomes (EQ-5D-5L domains and WHODAS domain scores):** cumulative odds ratio from a cumulative logit model.
- **Continuous outcomes (EQ-5D-5L VAS and WHODAS summary score):** mean difference from a linear mixed model.
- **Adherence proportion:** difference on the logit scale from a beta mixed model.

Shell tables of the planned comparative results are given at the end of this plan (@tbl-outcomes-analysis-results, @tbl-outcomes-analysis-results-all, and @tbl-additional-analyses-results).

# Analysis methods {#sec-analysis-methods}
<!-- github-issue-26: replace checklist placeholder with overview pointing to existing subsections. -->
We will analyse the primary and secondary outcomes with mixed-effects regression models that account for clustering at the hospital level and target cluster-specific intervention effects, matching the link function and outcome distribution to each outcome's data type; see @sec-effect-measures and the subsections below. We will use model-based standard errors, small-sample adjustments (including Kenward–Roger degrees-of-freedom corrections where relevant), and Wald-type confidence intervals, as specified under @sec-analysis-of-the-primary-outcome. We will fit the models by maximum likelihood with a Laplace approximation for the random effects, and choose the software implementation closer to analysis. We specify the covariate adjustment for the fully adjusted analysis under @sec-adjusted-analyses. Our model diagnostics will assess distributional assumptions, convergence, and, where relevant, the intercurrent-event strategy assumptions noted under @sec-intercurrent-events. If the preferred primary model does not converge, we will use the prespecified sequence of simpler models under @sec-analysis-of-the-primary-outcome.

## Analysis of the primary outcome {#sec-analysis-of-the-primary-outcome}
The primary outcome is in-hospital mortality within 30 days of arrival at the emergency department, which we will analyse as a dichotomous variable. We will estimate the odds ratio (OR) and the absolute risk difference (ARD) of death under ATLS^®^ compared with standard care.

<!-- sap-revision-todo B8 (comment id=20 AO; Round Two 27a): confirm hospital-level clustering, outcome-appropriate links, and cluster-specific intervention effect. -->
<!-- github-issue-13: CRT SAP reporting paper — time-averaged (common) SW intervention effect vs period-specific effects. -->
We will use a mixed-effects binomial regression model to estimate the effect of ATLS^®^ on mortality. We will estimate the OR using the logit link and the ARD using the identity link. These models account for clustering at the hospital level and give cluster-specific estimates of the intervention effect. Our primary estimand is a common intervention effect across periods of ATLS^®^ exposure — a time-averaged effect under a constant treatment-effect assumption — rather than separate period-specific effects after transition. For the primary analysis we will fit a prespecified sequence of models of decreasing complexity and report the most complex one that converges, moving to simpler models only if we meet convergence or numerical stability problems. We will then run the sensitivity analyses specified below. If the binomial model with the logit link converges but the corresponding model with the identity link does not, then we will report only an OR.

All models in this sequence will include a fixed effect for intervention exposure and adjust for time, although the parametrisation of time and the random effects structure may be simplified.

The model sequence is defined as follows:

1. **Primary model**: a model including fixed effects for period, modelled as a categorical variable with separate period effects for each batch, and a fixed effect for intervention exposure. Clustering will be accounted for by including a random intercept for cluster and a random cluster-by-period effect, both assumed to follow a normal distribution (@eq-primary-model).

$$
g\{\Pr(Y_{bkti} = 1)\} = \mu + \beta_{bt} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt}
$$ {#eq-primary-model}

2. **Simplified random effects structure**: a model including fixed effects for period and intervention exposure and a random intercept for cluster only, thereby removing the random cluster-by-period effect (@eq-simplified-random-effects-model).

$$
g\{\Pr(Y_{bkti} = 1)\} = \mu + \beta_{bt} + \theta X_{bkt} + \alpha_{bk}
$$ {#eq-simplified-random-effects-model}

<!-- sap-revision-todo B1 (comments id=191 KH, id=192 AO): state retained random effects and batch indexing for the shared period effects model. -->
3. **Model with shared period effects**: a model in which fixed period effects are shared across batches, with time adjusted for using a common period effect or calendar time modelled directly (e.g., as a continuous covariate or using splines). A random intercept for cluster and a random cluster-by-period effect are retained; clusters remain indexed by batch, but the fixed period effects are common across batches (@eq-shared-period-model).

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \beta_{t} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt}
$$ {#eq-shared-period-model}

4. **Combined simplified model**: a model that includes both a simplified random effects structure (random intercept for cluster only) and shared period effects. This represents the most parsimonious mixed-effects model considered acceptable for the primary analysis (@eq-combined-simplified-model).

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \beta_{t} + \theta X_{bkt} + \alpha_{bk}
$$ {#eq-combined-simplified-model}

Where:

- $g(\cdot)$: the link function, specified as either the logit link, $g(p) = \log\{p/(1-p)\}$, or the identity link, $g(p) = p$, depending on the estimand of interest.
- $\theta$: **the intervention effect**, represented as a fixed effect of intervention exposure (the primary estimand); under the logit link this represents the log-odds ratio for death associated with ATLS^®^ exposure compared with standard care, and under the identity link it represents the absolute risk difference.
- $\Pr(Y_{bkti} = 1)$: the probability of death for patient $i = 1, \dotsc, m_{bkt}$ in cluster $k = 1, \dotsc, `r clusters / batches`$ during period $t = 1, \dotsc, `r total.months - 1`$ in batch $b = 1, \dotsc, `r batches`$, giving `r clusters` clusters in total.
- $m_{bkt}$: the number of patients in cluster $k$ during period $t$ in batch $b$.
- $\mu$: the intercept, representing the value of the linear predictor for the reference levels of all fixed effects, with random effects equal to zero.
- $\beta_t$: the fixed effect of period $t$, shared across batches.
- $\beta_{bt}$: the fixed effect of period $t$ in batch $b$, accounting for a separate period effect for each batch, resulting in a total of `r batches` $\times$ (`r total.months - 1`) period effects.
- $X_{bkt}$: the intervention exposure indicator for cluster $k$ during period $t$ in batch $b$, with $X_{bkt} = 1$ for ATLS^®^ and $X_{bkt} = 0$ for standard care.
- $\alpha_{bk}$: a normally distributed random effect for cluster $k$ in batch $b$, representing between-cluster (hospital-level) variability in the outcome.
- $\gamma_{bkt}$: a normally distributed random effect for cluster $k$ in period $t$ in batch $b$, representing within-cluster variation over time and capturing cluster-specific deviations from the average period effect.

The distributional assumptions for the random effects are specified in @eq-distributional-assumptions.

$$
\begin{aligned}
\alpha_{bk} &\overset{\text{iid}}{\sim} \mathcal{N}(0,\sigma^2_{\alpha}), \\
\gamma_{bkt} &\overset{\text{iid}}{\sim} \mathcal{N}(0,\sigma^2_{\gamma}), \\
\alpha_{bk} &\perp \gamma_{bkt}.
\end{aligned}
$$ {#eq-distributional-assumptions}

<!-- github-issue-15: deduplicate B3/A1/B2 with overview and software sections. -->
<!-- sap-revision-todo B3 (comments id=21, id=22 AO; Round Two 27c/27d): model-based SEs, small-sample adjustments, Wald-type CIs. -->
We will use model-based standard errors and Wald-type confidence intervals. Because few clusters can inflate the type I error rate, we will apply a small-sample correction. We will use the Kenward–Roger correction [@grantham_evaluating_2022], but may revise that choice in light of methodological developments available closer to the end of patient inclusion. We may use a different correction for outcomes collected in the complete stepped-wedge design and in the incomplete nested staircase designs.

<!-- sap-revision-todo A1 (comments id=199 KH, id=200 AO); B4 (comment id=23 AO; Round Two 27b): general estimation framework is ML with Laplace; package/implementation not prespecified. -->
We will fit the mixed-effects models by maximum likelihood, using a Laplace approximation to integrate out the random effects; we do not prespecify the R package or numerical implementation, as noted under @sec-statistical-software.

<!-- sap-revision-todo B2 (comments id=24 MGW, id=25 AO; Round Two): state that diagnostics will check distributional assumptions and convergence. -->
We will use model diagnostics to check distributional assumptions and model convergence.

## Analysis of secondary outcomes {#sec-analysis-of-secondary-outcomes}
<!-- github-issue-15: keep only ordinal-specific diagnostics here; general diagnostics under primary. -->
<!-- sap-revision-todo B2 (comments id=24 MGW, id=25 AO; Round Two): diagnostics for secondary models, including proportional odds for ordinal outcomes. -->
<!-- sap-revision-todo C5 (comments id=288, id=289 KH; id=290, id=291 AO): nested staircase collection; period effects on global timeline, stated once here. -->
Unless otherwise stated, we will analyse secondary outcomes within the same mixed-effects framework as the primary outcome: the fixed effects for time and intervention exposure and the random-effects structure of the model selected for the primary analysis, with the link function and outcome distribution adapted to the scale of the outcome. Below, "the model selected for the primary analysis" means the most complex model in that sequence that converges; see @sec-analysis-of-the-primary-outcome. For the nested staircase outcomes (adherence, quality of life, and disability) we define period effects on the global study timeline rather than separately within each batch, and not all clusters contribute data in every period. For cumulative logit models our diagnostics will also assess the proportional-odds assumption.

### All-cause mortality within 24 hours, 30 days and three months of arrival at the emergency department {#sec-all-cause-mortality-within-24-hours-30-days-and-three-months-of-arrival-at-the-emergency-department}
We will analyse these outcomes as dichotomous variables using the mixed-effects binomial model selected for the primary analysis, with the logit and identity links, to estimate odds ratios and absolute risk differences.

### Quality of life within seven days of discharge, and at 30 days and three months of arrival at the emergency department, among survivors {#sec-quality-of-life-within-seven-days-of-discharge-and-at-30-days-and-three-months-of-arrival-at-the-emergency-department-among-survivors}
<!-- sap-revision-todo C6 (comments id=295, id=296 KH/AO; id=303 KH): clarify EQ-5D-5L yields ordinal domains plus VAS. -->
<!-- sap-revision-todo C7 (Anna email): standardise instrument name to EQ-5D-5L in SAP text. -->
We measure quality of life within the nested staircase design, using the official and validated translations of the EQ-5D-5L. It yields two outcome types: five ordinal domain scores (mobility, self-care, usual activities, pain/discomfort, and anxiety/depression), each rated on a Likert scale from 1 to 5, and a visual analogue scale (VAS) ranging from 0 to 100.

For each of the five ordinal domains, we will use a mixed-effects cumulative logit (proportional odds) model:

$$
\text{logit}\{\Pr(Y_{bkti} \leq j)\}
= \mu_j + \beta_{t} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt}
$$ {#eq-ordinal-model}

Where $\mu_j$ is the intercept for the $j$th category.

The VAS will be analysed as a continuous outcome using a linear mixed-effects model:

$$
\text{VAS}_{bkti}
= \mu + \beta_{t} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt} + \epsilon_{bkti},
\quad \epsilon_{bkti} \sim \mathcal{N}(0,\sigma^2)
$$ {#eq-linear-model}

### Disability within seven days of discharge, and at 30 days and three months of arrival at the emergency department, among survivors {#sec-disability-within-seven-days-of-discharge-and-at-30-days-and-three-months-of-arrival-at-the-emergency-department-among-survivors}
We measure disability within the nested staircase design using the WHO Disability Assessment Schedule 2.0 (WHODAS 2.0) [@ustun_measuring_2010], which assesses six domains of functioning: cognition, mobility, self-care, getting along, life activities, and participation. Each domain is rated on a Likert scale from 1 to 5, with 1 indicating no difficulties and 5 indicating extreme difficulties. We will analyse each domain separately using a mixed-effects cumulative logit model (@eq-ordinal-model). We will also calculate a WHODAS 2.0 summary score using the "complex scoring" method: we sum the item scores within each of the six domains, sum the six domain scores, and transform the total to a 0–100 scale. We will analyse the summary score using a linear mixed-effects model (@eq-linear-model).

### Return to work at 30 days and three months after arrival at the emergency department, among survivors {#sec-return-to-work-at-30-days-and-three-months-after-arrival-at-the-emergency-department-among-survivors}
<!-- github-issue-10: align RTW with binary OR and ARD (same as primary mortality). -->
<!-- github-issue-18: use converged primary-analysis model, not a hard-coded @eq-primary-model. -->
We will analyse return to work as a dichotomous outcome using the mixed-effects binomial model selected for the primary analysis, with the logit and identity links, to estimate odds ratios and absolute risk differences.

<!-- sap-revision-todo A2 (comments id=349, id=350 KH; id=351 AO; id=348 AO): replace Fine–Gray competing-risks LOS models with mixed-effects negative binomial count models; clarify purpose, random/period effects, and death/transfer handling. -->
<!-- sap-revision-todo C2 (comment id=348 AO): ED stay defined as hours to ED exit; exit routes enumerated; all routes contribute observed duration. -->

### Length of emergency department stay {#sec-length-of-emergency-department-stay}
We will analyse length of emergency department stay as a count outcome: the number of hours from arrival at the emergency department (ED) until ED exit. Exit may occur via ward admission, ICU admission, transfer to another hospital, death, or discharge home; all of these routes contribute the observed ED duration. We will estimate the effect of ATLS^®^ training on mean ED length of stay. We analyse mortality separately as a primary and secondary outcome rather than as a competing event.

<!-- github-issue-19: preferred LOS model with same convergence sequence as primary. -->
The preferred model is a mixed-effects negative binomial regression with a log link, including fixed effects for intervention exposure and batch-specific categorical period effects, a random intercept for cluster, and a random cluster-by-period effect (@eq-length-of-stay-model). If this model does not converge, we will simplify the random-effects and period structure following the same sequence of decreasing complexity as for the primary outcome. The intervention effect $\theta$ is the log rate ratio for ED length of stay associated with ATLS^®^ exposure compared with standard care; we will also report the corresponding multiplicative effect on the mean count (rate ratio).

$$
\log\{\mathbb{E}(Y_{bkti})\}
= \mu + \beta_{bt} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt}
$$ {#eq-length-of-stay-model}

Where $Y_{bkti}$ is the length-of-stay count for patient $i$ in cluster $k$ during period $t$ in batch $b$, and all other terms are defined as in the primary analysis model. The negative binomial model also estimates a dispersion parameter.

### Length of hospital stay {#sec-length-of-hospital-stay}
We will analyse length of hospital stay as a count outcome: the number of days from arrival at the emergency department to hospital discharge (or death in hospital). We will use the same preferred negative binomial specification and the same convergence sequence as for ED length of stay (@eq-length-of-stay-model).

### Intensive care unit admission {#sec-intensive-care-unit-admission}
<!-- github-issue-10: align ICU admission with binary OR and ARD (same as primary mortality). -->
We will analyse intensive care unit admission as a dichotomous outcome using the mixed-effects binomial model selected for the primary analysis, with the logit and identity links, to estimate odds ratios and absolute risk differences.

### Length of intensive care unit stay {#sec-length-of-intensive-care-unit-stay}
We will analyse length of intensive care unit stay as a count outcome among patients admitted to ICU: the number of days from ICU admission to ICU discharge (or death in ICU). We will use the same preferred negative binomial specification and the same convergence sequence as for the other length-of-stay outcomes (@eq-length-of-stay-model).

### Adherence to ATLS^®^ principles during initial patient resuscitation {#sec-adherence-to-atls-principles-during-initial-patient-resuscitation}
<!-- sap-revision-todo C3 (comment id=361 KH): checklist assessed in both standard-care and ATLS® periods. -->
<!-- sap-revision-todo C4 (comments id=61 KH, id=62 AO, id=63 MGW): express adherence as percentage; 14/14 = 100%; duplicate paragraph already removed. -->
We assess adherence to ATLS^®^ principles within the nested staircase design, in both the standard care and the ATLS^®^ periods, as the percentage of completed items on a 14-item checklist that covers the key steps of the ATLS^®^ primary survey, where all 14 steps completed is 100% adherence. We will rescale the percentage to a proportion on the 0–1 scale and analyse it with a mixed-effects beta regression model with a logit link:

$$
\text{logit}\{\mathbb{E}(Y_{bkti})\}
= \mu + \beta_{t} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt}
$$ {#eq-beta-model}

Where $\mathbb{E}(Y_{bkti})$ denotes the expected adherence proportion. The intervention effect $\theta$ represents the difference in expected adherence proportion on the logit (link) scale. Because the observed adherence proportion may take values of 0 or 1, values will be rescaled to lie in the open interval $(0,1)$ prior to fitting the beta regression model. A beta regression precision (dispersion) parameter will be estimated as part of the model.

<!--
*Question:** Should we adjust for calendar time because the last batch will be two years later than the first batch?

JK: Would this be instead of the beta_bt terms? That is, instead of separate period effects for each batch, just terms for each period would be included? (This is related to the question about 78 parameters in the model). The separate period effects for each batch will adjust for the calendar time. This is the model that makes the fewest assumptions about the period effects.

We've done a bit of work recently which shows that if there are separate period effects in each batch but instead shared period effects are included, this can lead to quite a lot of bias in the estimation of the treatment effect. This work also seems to indicate that when separate period effects are included but the period effects are actually shared, this doesn't really have any deleterious impacts on the inference for the treatment effect. (However note that parameterisation of time wasn't the focus of the work we've done).

So my recommendation would be for the primary analysis to be as currently specified (separate period effects for each batch); then if there are issues with fitting that model, try a model with calendar time (i.e. period effects shared when there is overlap between successive batches).

**Question:** Just a thought: keep in mind the complexity of the random effects structure; if it leads to convergence problems or overfitting, should we consider simplifying the random effects?

-->

## Correlations {#sec-correlations}
<!-- github-issue-26: move ICC/correlation reporting here from primary-analysis subsection (B5–B6). -->
<!-- sap-revision-todo B5 (comments id=26, id=27 AO; Round Two): report time-adjusted within-cluster correlations, variance components, and latent-scale correlations (protocol-aligned). -->
<!-- sap-revision-todo J6 (amended, sap-coherence-review R1-22): latent-scale reporting is additional for binary and ordinal outcomes. -->
We will report time-adjusted within-cluster correlations for all outcomes with 95% CIs, including the intra-cluster correlation (ICC) and within- and between-period correlations implied by the assumed correlation structures, together with all estimated variance components. For binary and ordinal outcomes we will additionally report these correlations on the latent scale; for continuous outcomes the latent and observed scales coincide.

<!-- sap-revision-todo B6 (comments id=28, id=29 AO; Round Two): ICC and related correlations derived from variance components; AR(1) rho from the AR(1) structure. -->
Intra-cluster and related correlation parameters will be derived from the estimated variance components of the fitted models. For models with an AR(1) within-cluster structure, the correlation parameter will be estimated directly from the AR(1) structure. We will present these parameters in a supplementary table (@tbl-correlation-parameters).

## Missing data {#sec-missing-data}
<!-- sap-revision-todo A5 (comment id=260 AO): 10% primary-outcome missingness chooses available-case pipeline vs MICE; do not mix complete-case adjusted analyses with covariate imputation. -->
We will present the frequency and percentage of missing data for all variables. How we handle missing data in the primary analysis will depend on the percentage of missing primary outcome data.

If less than 10% of patients have a missing primary outcome, we will use an available-case approach. The primary unadjusted analysis will include all patients with observed primary outcome data. We will restrict adjusted analyses to patients with complete data on the adjustment covariates, and repeat the unadjusted analysis in this complete-case population so that the unadjusted and adjusted results are directly comparable.

<!-- sap-revision-todo B7 (comments id=31 MGW, id=32 AO; Round Two): MI model reflects hospital clustering plus period and intervention exposure. -->
If 10% or more of patients have a missing primary outcome, we will use multiple imputation by chained equations (MICE) under a missing-at-random assumption, imputing the primary outcome, secondary outcomes as required for analysis, and the covariates included in the fully adjusted model, with a minimum of 20 imputations. The imputation model will reflect the hierarchical data structure, including clustering at the hospital level and design variables such as period and intervention exposure. We will then run the unadjusted and adjusted analyses on the imputed datasets, and also present the corresponding complete-case analyses for comparison. We will not impute secondary outcomes under the available-case approach.

If there is evidence that data are missing not at random, we will explore this in a sensitivity analysis.

<!--
Thoughts on missing data:

Check nature of missing data,
If MCAR, complete case analysis is generally acceptable, as there would be no systematic differences between those with and without missing data.
If MAR, then MICE is appropriate.
If MNAR, MICE may not fully address the missingness. In such cases, a sensitivity analysis exploring MNAR models might be useful (e.g., pattern mixture models or selection models).

Perhaps to strengthen the missing data plan, we could consider adding a few lines:

If there is evidence that the data is missing not at random (MNAR), we will explore the impact of this assumption using a sensitivity analysis (e.g., pattern mixture models or selection models) to assess how robust our findings are to different assumptions about the missing data mechanism. Additionally, we will perform diagnostic checks after multiple imputation to ensure the quality of the imputation process, including comparing distributions of observed and imputed data and checking convergence.

----

Let us discuss this further.

-->

## Sensitivity analyses {#sec-sensitivity-analyses}
<!-- github-issue-26: checklist covered by existing sensitivity subsections; note convergence fallback. -->
We will conduct sensitivity analyses to assess how robust the primary analysis results are to the modelling assumptions, the definition of intervention exposure, and the missing-data assumption, focusing throughout on the stability of the estimated intervention effect $\theta$. Unless otherwise stated, each sensitivity analysis uses two separate models, one with the logit link and one with the identity link. If a sensitivity model does not converge, we will report that and, where feasible, fit the next simpler random-effects or period specification, analogous to the primary-analysis convergence sequence.

We will explore whether models with more complex correlation structures fit the data better. We do not use them for the primary analysis because it is not well understood when they converge or how best to choose between plausible correlation structures.

### Models with autoregressive within-cluster correlation {#sec-models-with-autoregressive-within-cluster-correlation}
We will first explore whether allowing temporal correlation between outcomes within a cluster changes our inference about the intervention effect. Specifically, we will consider a discrete time-decay correlation structure by specifying a random cluster effect with an autoregressive structure of order one (AR(1)), described in @eq-ar1. This parameterisation assumes that correlations between cluster-level random effects decay geometrically with increasing temporal separation between periods.

<!-- sap-coherence-review (R1-15): the AR(1) subsection stated only the recursion, leaving the rest of the model unstated, and reused $\epsilon$ and $\sigma^2_\alpha$ for the innovation. -->
The AR(1) process replaces the random intercept $\alpha_{bk}$ of the primary analysis model; all other terms, including the cluster-by-period random effect $\gamma_{bkt}$, are unchanged.

$$
\alpha_{bk,t} = \rho \alpha_{bk,t-1} + \eta_{bkt}, \quad \eta_{bkt} \sim \mathcal{N}(0, \sigma^2_\eta)
$$ {#eq-ar1}

Where:

- $\alpha_{bk,t}$: the random effect for cluster $k$ in period $t$ in batch $b$.
- $\rho$: the correlation between adjacent periods within the same cluster.
- $\eta_{bkt}$: a normally distributed innovation with variance $\sigma^2_\eta$, which is distinct from the marginal cluster variance $\sigma^2_{\alpha}$ in @eq-distributional-assumptions.

This AR(1) formulation differs slightly from alternative parameterisations in which correlations are specified through a joint covariance matrix. Both capture decaying temporal correlation, but $\rho$ is not directly equivalent to the parameters of those specifications and should not be interpreted interchangeably.

<!-- sap-revision-todo A6 (comments id=212 KH, id=213 AO): removed "Models with batch-specific secular trends" sensitivity; primary model already uses fixed batch-specific period effects β_bt (protocol-aligned); shared period effects remain in the convergence sequence. -->

### Models with random cluster by intervention effects {#sec-models-with-random-cluster-by-intervention-effects}
<!-- sap-revision-todo A3 (comments id=220 KH, id=221 AO): dropped the over-complex three-way correlated RE section; retained a simplified random intervention slope with non-zero Cov(α,u) only (γ independent) to adhere to the protocol while addressing reviewer complexity concerns. -->
To assess sensitivity to the assumption of a common intervention effect across clusters, we will extend the primary analysis model to allow the intervention effect to vary by cluster. Specifically, we will include a random cluster-level slope for the intervention exposure, with a non-zero covariance between this slope and the cluster random intercept (as anticipated in the protocol). The cluster-by-period random effect remains independent of both (@eq-model-random-cluster-intervention).

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \beta_{bt} + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt} + u_{bk} X_{bkt}
$$ {#eq-model-random-cluster-intervention}

Where:

- $u_{bk}$: a random effect representing cluster-specific deviations from the average intervention effect.
- The cluster intercept and intervention slope are jointly normally distributed,
$$
\begin{pmatrix}
\alpha_{bk} \\
u_{bk}
\end{pmatrix}
\sim
\mathcal{N}\!\left(
\begin{pmatrix}
0 \\
0
\end{pmatrix},
\begin{pmatrix}
\sigma^2_{\alpha} & \sigma_{\alpha u} \\
\sigma_{\alpha u} & \sigma^2_{u}
\end{pmatrix}
\right),
$$
with $\sigma_{\alpha u} \neq 0$ allowed.
- $\gamma_{bkt} \sim \mathcal{N}(0,\sigma^2_{\gamma})$ is independent of $(\alpha_{bk}, u_{bk})$.

### Models with time modelled with a spline function {#sec-models-with-time-modelled-with-a-spline-function}
We will further explore the potential for a time-varying treatment effect [@kenny_analysis_2022]. To explore if the fixed period effect is both parsimonious and adequate to represent the extent of any underlying secular trend, we will model calendar time using a shared natural cubic spline basis across all batches with knots at the equally spaced time points 3, 6 and 9. This will result in five spline basis functions, because the natural cubic splines are modelled with three degrees of freedom but are constrained to be linear before the first and after the last knot. The model is specified in @eq-spline-model.

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \sum_{j=1}^{5} \beta_j S_j(t, \{3,6,9\}) + \theta X_{bkt} + \alpha_{bk} + \gamma_{bkt}
$$ {#eq-spline-model}

Where:

- $t$: calendar time (period index) common to all batches.
- $S_j(t, \{3,6,9\})$: the $j$th natural cubic spline basis function, shared across batches.
- $\beta_j$: the coefficient for the $j$th spline basis function.

### Models exploring lag and weaning effects {#sec-models-exploring-lag-and-weaning-effects}
We will also extend the primary analysis model with an interaction between treatment and the number of periods since first treated, to examine whether longer exposure to the intervention changes outcomes. This lets us model lag effects, where the intervention takes time to become embedded in the culture before it has an effect, and weaning effects, where its effect fades. This type of analysis attempts to disentangle how some clusters end up having a long exposure to the intervention and others have a much shorter exposure time (@eq-lag-weaning-model).

$$
g\{\Pr(Y_{bkti} = 1)\}
= \mu + \beta_{bt} + \theta X_{bkt}
+ \theta_{\text{int}} X_{bkt} \times T_{bkt}
+ \alpha_{bk} + \gamma_{bkt}
$$ {#eq-lag-weaning-model}

Where $\theta_{\text{int}}$ represents the interaction between intervention exposure and time since first treated, and $T_{bkt}$ is the number of periods since first exposure.

<!-- sap-coherence-review (R1-02, R2-06): the actual-transition-date sensitivity analysis was promised under Analysis sets, Adherence to the intervention and Intercurrent events but specified nowhere, and @tbl-additional-analyses-results already reserves rows for it. -->
### Exposure defined by the actual transition date {#sec-exposure-defined-by-the-actual-transition-date}
Where a cluster's actual transition date differs from its scheduled date by more than four weeks, we will refit the primary analysis model with the intervention exposure indicator $X_{bkt}$ defined by the actual rather than the scheduled transition date. We exclude observations in the month of the actual transition, as we do for the scheduled transition, and leave all other model terms unchanged. We will report the number of clusters for which we redefine exposure.

## Adjusted analyses {#sec-adjusted-analyses}
In the fully adjusted analysis we will additionally adjust for<!--[^additional-covariates]--> (expected % missing data in parentheses):

-   Age (<5%)
-   Sex (<5%)
-   Systolic blood pressure (25%)
-   Glasgow Coma Scale (25%)
-   Injury Severity Score (10%)
-   Mechanism of injury (10%)

These are known individual-level prognostic factors for the primary outcome. We will add these covariates as fixed effects to the model selected for the primary analysis. We will model the continuous covariates assuming linear effects.

<!--
[^additional-covariates]: **Questions:** 1. Should we adjust for calendar time because the last batch will be two years later than the first batch? 2. Should we add it as covariate in the list above? 3. Should it be continuous (e.g., months since the study started) or categorical calendar time (if you expect non-linear trends or stepwise changes over years)? 4. Should we analyse Time Since Admission? Depending on the nature of the trauma, the time since patient admission might also influence outcomes, especially if the timing of interventions or care decisions plays a critical role. Eg. time from admission to surgery.
-->

## Subgroup analyses {#sec-subgroup-analyses}
<!-- github-issue-26: remove checklist placeholder; existing F1/F2 text covers methods. -->
<!-- sap-revision-todo F1 (comments id=33 MGW, id=34 AO; id=368 KH): subgroup analyses of primary outcome via interactions. -->
<!-- sap-revision-todo F2 (comments id=243, id=246 KH): clarify subgroup coding; cluster size categorical; keep six subgroups. -->
<!-- github-issue-16 github-issue-17: subgroup-specific effects from interaction model; use converged primary model. -->
We will perform the following subgroup analyses of the primary outcome:

<!-- sap-revision-todo I3 (comments id=238, id=239): number of states depends on final clusters. -->
-  geographical region, defined using the state in which the participating hospital is located, with each state as a separate category; the number of states depends on the final set of participating clusters and cannot be prespecified;
-  age groups, defined as older adolescents (15–19 years), young adults (20–24 years), adults (25–59 years), and older adults (60 years and older) [@Diaz2021];
-  sex, using the levels male and female;
-  clinical cohorts, defined as blunt multisystem trauma, penetrating trauma, and severe isolated traumatic brain injury, with modification to avoid overlap between the cohorts [@hornor_quality_2018]; this subgroup analysis will include only patients classified into one of these three cohorts;
-  patients with major trauma, defined as patients with an injury severity score (ISS) of 16 or higher versus those with ISS below 16;
-  cluster size, defined as the monthly volume of eligible patients (as used for covariate-constrained randomisation) and categorised as small (fewer than 12 patients per month), medium (12–20), or large (more than 20)

As in the protocol, we treat geographical region as our primary subgroup and the rest as exploratory.

Each subgroup analysis will use the mixed-effects binomial model selected for the primary analysis, with the logit and identity links. We will add the subgroup variable and the interaction between the subgroup variable and intervention exposure as fixed effects. From that model we will obtain the intervention effect within each subgroup level, as odds ratios and absolute risk differences, by combining the main intervention effect with the corresponding interaction term, and report each with a 95% CI. We will also report the interaction p-value as a test of heterogeneity across subgroup levels. We will assume no interaction between subgroups and secular time trends, which may be evaluated in future analyses if indicated.

## Safety events {#sec-safety-events}
<!-- sap-revision-todo J9 (GitHub #8): final-analysis safety event shell table; protocol-aligned definitions. -->
We will summarise safety events, as defined in the trial protocol, as counts and percentages of patients with at least one recorded event, stratified by before and after ATLS^®^ training and overall (@tbl-safety-events).

## Statistical software {#sec-statistical-software}
<!-- github-issue-26: expand software note for checklist. -->
<!-- github-issue-13: CRT SAP reporting paper — any mention of analysis code counts toward code-reporting item. -->
We will use the R Statistical Software for all analyses [@R]. We will select the packages for mixed-effects modelling, small-sample corrections, multiple imputation, and reporting tables closer to the time of analysis, based on convergence behaviour and available methodology, and document them in the analysis code and final report. We will archive the analysis code used for the final analysis with the trial materials.

## Presentation of analysis results {#sec-presentation-of-analysis-results}
<!-- sap-revision-todo J2: gtsummary tbl_regression-style key/all shells; shared descriptive section headers and label style. -->
<!-- github-issue-34: outcome names as nested headers; effect measures indented with modify_indent. -->
We will report intervention-effect estimates for the primary and secondary outcomes with 95% CIs and p-values, using the effect measures specified for each outcome (@tbl-outcomes-analysis-results for key analyses, @tbl-outcomes-analysis-results-all for all analyses).

<!-- sap-revision-todo J3: shell for sensitivity, fully adjusted, and subgroup analyses of the primary outcome. -->
<!-- github-issue-34: analysis names as nested headers; OR/ARD indented with modify_indent. -->
We will report the sensitivity, fully adjusted, and subgroup analyses of the primary outcome in the same style (@tbl-additional-analyses-results), and present the subgroup-specific intervention effects in a forest plot with 95% CIs for the odds ratio and absolute risk difference (@fig-subgroup-forest-plot).

<!-- sap-revision-todo J10 (GitHub #9): forest plot shell for primary-outcome subgroup analyses. -->

<!-- github-issue-33: move reporting shells to end matter; keep design figures and outcomes summary inline. -->
# Shell figures and tables {#sec-shell-figures-and-tables}
## CONSORT diagrams {#sec-consort-diagrams}
<!-- github-issue-33: keep the three CONSORT shells consecutive; pin with fig-pos H so floats do not interleave tables. -->

```{r}
#| echo: false
#| label: fig-consort-clusters
#| fig.cap: "Cluster-level CONSORT diagram shell. Placeholders cover clusters assessed and randomised, losses and exclusions after randomisation with reasons, and clusters included in the primary analysis, by sequence and overall. The flowchart pools all six batches."
#| fig.align: "center"
#| fig-pos: H
#| out-width: "100%"
#| message: FALSE
#| warning: FALSE

knitr::include_graphics(
  create_cluster_consort_diagram(
    sequences = sequences,
    batches = batches,
    total.months = total.months,
    min.standard.care.months = min.standard.care.months,
    transition.months = transition.months,
    return.figure = FALSE,
    device = "png"
  )
)
```

```{r}
#| echo: false
#| label: fig-consort-patients
#| fig.cap: "Patient-level CONSORT diagram shell. Placeholders cover patients entered, losses and exclusions from the primary analysis with reasons, and patients analysed, by sequence, before and after ATLS^®^ training, and overall. The flowchart pools all six batches."
#| fig.align: "center"
#| fig-pos: H
#| out-width: "100%"
#| message: FALSE
#| warning: FALSE

knitr::include_graphics(
  create_patient_consort_diagram(
    sequences = sequences,
    batches = batches,
    total.months = total.months,
    min.standard.care.months = min.standard.care.months,
    transition.months = transition.months,
    return.figure = FALSE,
    device = "png"
  )
)
```

```{r}
#| echo: false
#| label: fig-consort-nested-staircase
#| fig.cap: "Nested-staircase CONSORT diagram shell. Placeholders cover patients in staircase periods, the shift-stratified nested sample, losses and exclusions with reasons, and patients analysed for nested-staircase secondary outcomes, by sequence, before and after ATLS^®^ training, and overall. Coloured strip cells mark staircase months; grey cells are outside nested data collection. The flowchart pools all six batches."
#| fig.align: "center"
#| fig-pos: H
#| out-width: "100%"
#| message: FALSE
#| warning: FALSE

knitr::include_graphics(
  create_nested_staircase_consort_diagram(
    sequences = sequences,
    batches = batches,
    total.months = total.months,
    min.standard.care.months = min.standard.care.months,
    transition.months = transition.months,
    staircase.months = staircase.months,
    return.figure = FALSE,
    device = "png"
  )
)
```

<!-- sap-coherence-review: this subsection holds the descriptive and supplementary shells as well as the analysis-results ones, so the heading named only part of its contents. -->
## Descriptive, analysis and supplementary shells {#sec-descriptive-analysis-and-supplementary-shells}
```{r}
#| label: tbl-cluster-characteristics
#| tbl-cap: "Cluster characteristics"
#| echo: false
#| message: false
#| warning: false
#| results: asis
create_cluster_characteristics_table()
```

```{r}
#| label: tbl-patient-characteristics
#| tbl-cap: "Key patient characteristics"
#| echo: false
#| message: false
#| warning: false
#| results: asis
create_patient_characteristics_table()
```

```{r}
#| label: tbl-intercurrent-events
#| tbl-cap: "Intercurrent events by outcome"
#| echo: false
#| message: false
#| warning: false
#| results: asis
create_intercurrent_events_table()
```

```{r}
#| label: tbl-outcomes-descriptive
#| tbl-cap: "Descriptive summaries of key outcomes"
#| echo: false
#| message: false
#| warning: false
#| results: asis
create_outcomes_descriptive_table()
```

```{r}
#| label: tbl-safety-events
#| tbl-cap: "Safety events"
#| echo: false
#| message: false
#| warning: false
#| results: asis
create_safety_events_table()
```

```{r}
#| label: tbl-outcomes-analysis-results
#| tbl-cap: "Key primary and secondary outcome analysis results"
#| echo: false
#| message: false
#| warning: false
#| results: asis
create_outcomes_analysis_results_table()
```

```{r}
#| label: tbl-additional-analyses-results
#| tbl-cap: "Sensitivity, fully adjusted, and subgroup analysis results for the primary outcome"
#| echo: false
#| message: false
#| warning: false
#| results: asis
create_additional_analyses_results_table()
```

```{r}
#| label: fig-subgroup-forest-plot
#| fig.cap: "Forest plot of subgroup-specific intervention effects on in-hospital mortality within 30 days. Estimates and confidence intervals are placeholders to be completed at analysis."
#| fig.align: "center"
#| echo: false
#| message: false
#| warning: false

knitr::include_graphics(
  create_subgroup_forest_plot(
    return.figure = FALSE,
    device = "png"
  )
)
```

```{r}
#| label: tbl-patient-characteristics-all
#| tbl-cap: "All patient characteristics"
#| echo: false
#| message: false
#| warning: false
#| results: asis
create_patient_characteristics_table(all = TRUE)
```

```{r}
#| label: tbl-outcomes-descriptive-all
#| tbl-cap: "Descriptive summaries of all outcomes"
#| echo: false
#| message: false
#| warning: false
#| results: asis
create_outcomes_descriptive_table(all = TRUE)
```

```{r}
#| label: tbl-outcomes-analysis-results-all
#| tbl-cap: "All primary and secondary outcome analysis results"
#| echo: false
#| message: false
#| warning: false
#| results: asis
create_outcomes_analysis_results_table(all = TRUE)
```

```{r}
#| label: tbl-correlation-parameters
#| tbl-cap: "Within-cluster correlation parameters for all outcomes"
#| echo: false
#| message: false
#| warning: false
#| results: asis
create_correlation_parameters_shell_table(all = TRUE)
```

# References {#sec-references .numbered}
::: {#refs}
:::
